\documentclass[11pt,openany]{book}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{listings}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{fancyvrb}
\usepackage{microtype}
\usepackage{parskip}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=blue!60!black,
  citecolor=black,
}

\definecolor{codebg}{RGB}{248,248,248}
\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebg},
  frame=single,
  breaklines=true,
  keepspaces=true,
  columns=flexible,
}

\newtheorem{finding}{Finding}[chapter]
\newtheorem{retraction}{Retraction}[chapter]
\newtheorem{ceiling}{Ceiling}[chapter]

\title{\vspace{-2em}
  The Architectural Dichotomy and the Context Ceiling\\
  \vspace{0.3em}
  \large{Session 12 Findings from the Heinrich Forensics Toolkit}\\
  \vspace{0.2em}
  \normalsize{Forensic decomposition of the causal-bank family, the discovery
              that ``speed arm'' is two architectures with different prediction
              machinery, and why the 1.0 bpb target is a context-mechanism
              problem, not a scale problem.}
}
\author{Heinrich Measurement System\\
\textit{With chronohorn and decepticons}\\
April 18--19, 2026}
\date{}

\begin{document}
\frontmatter
\maketitle

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

We document Session 12 of the Heinrich measurement program: a session built
on session-11's discovery of the MDL-optimal two-band substrate partition, and
one that ended by overturning several early interpretations of that finding.
Three threads run through the session.

First, \emph{architectural}: the Heinrich library contains two distinct
causal-bank families that had been lumped together as ``the speed arm.'' They
are \emph{substrate-primary} (local path disabled, adaptive substrate carries
the prediction) and \emph{local-primary} (local convolution dominates the
prediction, substrate is a small augment). The two families have different
bpb ceilings, different effective-context behavior, and respond to different
architectural levers. Every session-11 finding that mixed them needs revisiting.

Second, \emph{mechanistic}: the substrate-primary family has a structural
context ceiling around 16 bytes regardless of scan half-life. Increasing the
EMA half-life from 2 to 512 moves the effective context only from 8 to 16
bytes---256$\times$ more theoretical memory buys 2$\times$ more actual
context. The saturation bpb is roughly 1.78, corresponding to byte-level
$n$-gram entropy at $n\approx8$. Neither longer training, more modes, nor
richer $\omega$ parameterization breaks through this ceiling in the current
codebase. Hash memory, despite session-11's plumbing fix, does not extend it
either. \emph{Experiment A from the chronohorn planning deck---``substrate
with longer half-life''---is already in the library and already negative.}

Third, \emph{methodological}: the session-11 ``MDL-optimal two-band partition''
finding was a mid-training snapshot, not a stable architectural feature.
Extending the same model (M4 $\to$ W2-B) from 20k to 50k training steps
shows the partition initially intensifies, then leaks into the content band.
The binary \texttt{two\_band\_partition} flag from the session-11-pass tooling
was too strict and was replaced with a continuous \texttt{partition\_score}
and four-way categorical verdict (\texttt{absent}, \texttt{leaky},
\texttt{partial}, \texttt{partitioned}). The EffDim metric still undercounts
working dimensionality on any model with a two-band partition---a class that
now appears to cover the entire pos-signal-enabled lasso/cr family.

The central claim is narrow but consequential: the 0.8 bpb gap between the
current byte-level frontier (1.785) and the 1.0 bpb target is almost entirely
a \emph{context-mechanism} problem. The substrate mechanism cannot use
information more than 16 bytes past. Scaling the mechanism doesn't help; it
needs replacement.

The untested experiment that remains: \texttt{substrate\_mode=gated\_delta} with
\texttt{local\_scale=0.0}. Every existing gated-delta model in the library runs
with \texttt{local\_scale=0.25}, and local conv dominates, hiding what gated-delta
would do alone. This is the single-flag pilot that determines whether
Mamba-class selective gating escapes the substrate ceiling.

\tableofcontents

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This book continues the session record begun with \texttt{session11\_book.tex}.
Where session 11 was an empirical survey and plumbing-hardening pass, session
12 is a forensic-decomposition session. The inputs were mostly existing
model checkpoints in the \texttt{/Volumes/Sharts/heinrich} library; the new
tooling was small; the compute spent on training was zero. Nearly every
finding here is the result of reading state off of already-trained weights or
running inference-time experiments.

The session structure reflects the question-answer cadence chronohorn and
Heinrich developed through the night: chronohorn asks five sharp questions,
Heinrich answers them with forensic measurements, both sides then refine
understanding, and chronohorn asks five more. We retain that shape here
because many of the findings are only legible as responses to specific
questions. The chapters are organized by question-theme rather than
strictly chronologically.

Where we disagree with session-11 claims, we say so explicitly in
retraction boxes. Where an interpretation we held at the start of session 12
turned out to be wrong by session end, we document both the original and the
correction. The goal is a record that a future reader can walk into cold and
trust.

\mainmatter

\part{The Architectural Dichotomy}

\chapter{Substrate-primary vs local-primary}
\label{ch:dichotomy}

\section{The two families}

The Heinrich library and the session-11 narrative used the term ``speed arm''
as if it referred to a single architectural family. Session 12 establishes
that it refers to two families sharing tokens-per-second-per-parameter
characteristics but doing prediction via structurally different machinery.

\begin{finding}[Two-family architectural dichotomy]
Models in the causal-bank speed arm decompose into:
\begin{itemize}
  \item \textbf{Substrate-primary}: \texttt{local\_scale=0.0}; the adaptive
  substrate plus learnable-$\omega$ carries essentially all of the prediction.
  Ablating the local conv at inference has zero bpb effect.
  \item \textbf{Local-primary}: \texttt{local\_scale>0} (typically 0.25); the
  local convolution is the primary predictor, and the substrate provides a
  small additive augment. Ablating the local conv at inference drives bpb
  from $\sim 2.0$ to $\sim 4.9$---the substrate alone is insufficient.
\end{itemize}
\end{finding}

Both families plateau in the $1.8$--$2.0$ bpb range on byte-level FineWeb-10B
at speed-arm scales, but they do so via mechanistically different prediction.
The dichotomy was invisible at the bpb level; it was only visible in
ablation-at-inference.

\section{Diagnostic: local-ablation at inference}

The cleanest test is to monkey-patch \texttt{\_local\_logits} on the model to
return zeros, then measure the bpb change on a held-out sequence. A
substrate-primary model has zero bpb change (local wasn't used). A
local-primary model shows a catastrophic jump (local was load-bearing).

\begin{longtable}{lrrr}
\toprule
model & bpb (default) & bpb (local ablated) & $\Delta$ \\
\midrule
byte-hrr-learnable-s8-50k           & 1.814 & 1.814 & 0.000 \\
scan-ablation-s8-50k                & 1.797 & 1.797 & 0.000 \\
mut-hashmem-fixed-20k               & 2.110 & 2.110 & 0.000 \\
\midrule
byte-lasso-pos-b2-s12-50k           & 2.022 & 4.890 & +2.868 \\
cr-s8-scandead-20k                  & 2.074 & 4.890 & +2.816 \\
\bottomrule
\end{longtable}

The top three are substrate-primary, confirmed by \texttt{local\_scale=0.0}
in their configs and by the complete invariance to local ablation. The
bottom two are local-primary, confirmed by the catastrophic jump.

\section{Why this matters}

Every claim in session-11 and early session-12 notes that compares bpb across
``speed arm'' models is mixing apples and oranges. Three examples:

\begin{enumerate}
  \item ``Scan-ablation beats learnable-s8 by 0.03 bpb'' reads as ``scan-dead is
  a small win.'' But scan-ablation and learnable-s8 are \emph{both}
  substrate-primary, so the comparison is valid and the 0.03 bpb difference
  comes from half-life choice (2 vs 512).
  \item ``Byte-lasso-pos-b2 at s12 achieves 2.02 bpb, worse than scan-ablation''
  reads as a failure of lasso. But byte-lasso-pos-b2 is local-primary; its
  substrate contributes almost nothing to prediction. The comparison is
  actually between ``sophisticated substrate over $\omega$-projected bytes''
  (scan-ablation) and ``local conv over bytes plus a small substrate
  coherence augment'' (lasso-pos-b2). They are different architectures.
  \item ``Longer half-life gives better bpb'' was inferred from local-primary
  models where local dominates; the substrate-mode half-life setting was
  decorative.
\end{enumerate}

The dichotomy recasts what the 1 bpb target looks like from each family:

\begin{itemize}
  \item Local-primary: context ceiling = \texttt{local\_window} + small EMA.
  At \texttt{local\_window=8}, the effective window is $\sim 16$ bytes and
  saturation sits near 2.0 bpb. At \texttt{local\_window=32}, projected
  saturation is $\sim 1.5$ bpb (still bounded below transformer).
  \item Substrate-primary: ceiling depends on whether the substrate
  mechanism can propagate long-range information. As Chapter 2 will show,
  the current EMA-based substrate cannot; it caps around 16-byte effective
  context and 1.78 bpb.
\end{itemize}

\part{The Context Ceiling}

\chapter{Effective context by prefix corruption}
\label{ch:context-ceiling}

\section{Methodology}

The prefix-corruption test measures how much of a prediction's quality
depends on real versus random prefix bytes. For a sequence of length
$L$, prepend $N$ random bytes and append $M = L - N$ real bytes from the
validation corpus. Measure cross-entropy at the final position as a function
of $M$. The ``knee'' is the smallest $M$ at which bpb first approaches its
saturation value.

An alternative measurement is the per-position bpb curve on a
fully-real-prefix sequence. A position at index $k$ has access to $k$ real
prior bytes. The bpb at position $k$ as a function of $k$ traces the
effective-context curve directly, giving $L-1$ samples per trial rather than
one.

Both methods produce the same knee in practice. The per-position method has
tighter noise bars and was used in session 12 for the tighter comparisons.

\section{The core measurement}

\begin{longtable}{lrrr}
\toprule
model & knee (bytes) & saturation bpb & config \\
\midrule
scan-ablation-s8-50k        & 8    & 1.80 & hl=2,   ls=0.0 \\
byte-hrr-learnable-s8-50k   & 16   & 1.78 & hl=512, ls=0.0 \\
byte-hrr-learnable-s12-50k  & 16   & 1.78 & hl=512, ls=0.0 \\
mut-hashmem-fixed-20k       & 16   & 2.09 & hl=512, ls=0.0 \\
mut-control-mlp-20k         & 16   & 2.08 & hl=512, ls=0.0 \\
\midrule
byte-lasso-pos-b2-s12-50k   & 16   & 2.02 & hl=512, ls=0.25, local-primary \\
cr-s8-scandead-20k          & 4    & 2.05 & hl=2,   ls=0.25, local-primary \\
\bottomrule
\end{longtable}

(hl = linear\_half\_life\_max; ls = local\_scale.)

Two observations immediately:

\begin{ceiling}[The EMA scan is structurally broken for long context]
The substrate-primary models at \texttt{hl=512} all have an effective
context of 16 bytes. Increasing \texttt{half\_life\_max} from 2 to 512
(a $256\times$ change in theoretical memory horizon) moves the knee from
8 to 16 bytes ($2\times$). Saturation bpb drops by $0.02$. The
substrate's EMA does not propagate long-range dependency into the readout
in any way that training exploits.
\end{ceiling}

\begin{finding}[Hash memory does not extend context]
\texttt{mut-hashmem-fixed-20k}, the post-fix hash-memory variant that
actually trains the hash module (session-11 book confirmed this), has the
same 16-byte knee and same 2.09 bpb saturation as its hash-less sibling
\texttt{mut-control-mlp-20k} (2.08 saturation). Retrieved context is either
too noisy or too weakly-gradient-connected to the prediction.
\end{finding}

The apparent ``speed arm ceiling at 1.8 bpb'' is actually the English byte
$n$-gram entropy at $n\approx 8$. Current architectures are sophisticated
$n$-gram predictors, not long-context models.

\section{Why EMA fails at long context}

The EMA update rule is:
\[
    h_t = \alpha \odot h_{t-1} + (1 - \alpha) \odot W_{\mathrm{in}} x_t, \quad
    \alpha = \exp(-\ln 2 / \tau)
\]
At half-life $\tau=512$, $\alpha \approx 0.9986$. The state at position $t$
theoretically retains a small but nonzero contribution from positions up to
several hundred steps past. But that retention is additive and undifferentiated:
every past token is blurred into a single running average. The readout has no
way to tell what happened 30 tokens ago versus what happened 3 tokens ago.

The relevant comparison: in Mamba-class models, the update gate is
\emph{input-dependent}, and the state accumulation is selective rather
than summary. A selective-gated model can choose to remember one specific
past event and discard the intervening ones. The EMA cannot.

This is what it means for the EMA to be ``structurally broken for long
context.'' It's not that the memory horizon is too short. It's that the
integration is lossy in a way training cannot repair.

\chapter{Substrate-ablation tests}
\label{ch:substrate-ablation}

\section{The substrate is doing real work}

The feedforward hypothesis---that scan-ablation is a scan-dressed
feedforward model whose prediction comes entirely from per-token
$\omega$-modulation---can be tested directly by zeroing the substrate state
contribution to the readout features.

\begin{longtable}{lr}
\toprule
condition & bpb \\
\midrule
full (normal inference) & 1.797 \\
substrate-modes zeroed (only $x_{\mathrm{embed}}$ tail passes) & 6.651 \\
$x_{\mathrm{embed}}$-tail zeroed (only substrate modes pass) & 1.850 \\
seq\_len$=1$ (no prior context) & 3.820 \\
\bottomrule
\end{longtable}

\begin{finding}[Substrate carries the prediction on substrate-primary models]
On scan-ablation-s8-50k, zeroing the substrate contribution to readout
features drives bpb from 1.80 to 6.65. Zeroing the \texttt{x\_embed} tail
alone only moves bpb from 1.80 to 1.85. The substrate is responsible for
essentially all the prediction quality; the concat-tail contribution is
decorative.
\end{finding}

The seq\_len$=1$ measurement is the cleanest feedforward test. With no
prior context, the model predicts 3.82 bpb---bigram-quality on English
bytes. The gap from 3.82 (1-byte context, feedforward) to 1.80 (full context,
normal) is 2.02 bpb. That is the substrate's context contribution.

The seq\_len$=1$ measurement also falsifies the feedforward hypothesis:
if scan-ablation were feedforward, bpb would be unchanged at 1.80 regardless
of context length. It is not. The substrate propagates information across
roughly 8--16 positions and that propagation accounts for 2 bpb of
improvement.

\section{Substrate dimensionality sensitivity}

The $\omega$ projection was found in session 11 to have effective rank
$\approx 28$ out of 256 input dimensions. A natural hypothesis: if the
projection is low-rank, the substrate itself may also be over-parameterized.
Zero out the modes beyond rank-$K$ and measure bpb:

\begin{longtable}{lrr}
\toprule
$K$ & bpb & $\Delta$ vs full (2048 modes) \\
\midrule
2048 (full) & 1.797 & --- \\
1024        & 1.797 & +0.000 \\
512         & 1.813 & +0.017 \\
256         & 2.009 & +0.212 \\
128         & 2.511 & +0.714 \\
64          & 3.053 & +1.256 \\
32          & 3.533 & +1.736 \\
16          & 3.965 & +2.168 \\
\bottomrule
\end{longtable}

\begin{finding}[Substrate supports 50\% free dimensionality reduction, no more]
Truncating scan-ablation-s8-50k's substrate to $K=1024$ modes at inference
produces identical bpb (1.797). Truncating to $K=512$ costs 0.017 bpb.
Below $K=256$ the model degrades sharply. This is not the $10\times$
efficiency lever the low-rank $\omega$ finding suggested; it is a
straightforward $2\times$ width reduction with no penalty.
\end{finding}

The $\omega$ projection's low rank (28) refers to the \emph{input} side of
the projection, not the output side. The output (substrate modes) uses
roughly 500--1000 dimensions genuinely, not 28. Future scaling should
budget $\sim$1024 modes as the minimum viable width.

\section{$\omega$ encodes byte-category structure}

A companion analysis: per-byte $\omega$ vectors (running all 256 bytes as
single-token inputs and extracting the captured $\omega$ tensor) form a
$256 \times 2048$ matrix. PCA of this matrix reveals what $\omega$ learns.

\begin{longtable}{lrr}
\toprule
metric & value & note \\
\midrule
effective rank of per-byte $\omega$ & 27.5 & (out of 256, low-rank) \\
top-10 PCs variance & 63.9\% & \\
top-32 PCs variance & 94.7\% & \\
\midrule
byte-category classification (top-10 PCs) & & (nearest centroid) \\
\quad digit & 100\% & (10 classes, perfect) \\
\quad whitespace & 100\% & \\
\quad punct & 84\% & \\
\quad letter & 73\% & \\
\quad control & 72\% & \\
\quad non-ASCII & 53\% & (chance) \\
\bottomrule
\end{longtable}

Nearest-neighbor pairs in the PC space:
\texttt{a}$\leftrightarrow$\texttt{A}, \texttt{e}$\leftrightarrow$\texttt{E},
\texttt{z}$\leftrightarrow$\texttt{Z}, \texttt{0}$\leftrightarrow$\texttt{1}.
$\omega$ groups case pairs and digit adjacency, both of which are
linguistically meaningful byte-category clusters in English.

\begin{finding}[$\omega$ is a sophisticated byte-category embedding, not a
lookup or an $n$-gram predictor]
The per-byte $\omega$ projection on scan-ablation-s8-50k has effective rank
$\approx 28$ and achieves 100\% classification accuracy on digits and
whitespace, 73--84\% on letter/punctuation sub-categories, and chance on
non-ASCII bytes. Case pairs are co-located in PC space. Since scan-ablation
has dead EMA (half-life 2), $\omega$ cannot encode $n$-gram context from
past tokens; what it learns is \emph{per-token} byte-category structure
with enough linguistic information to distinguish ASCII sub-classes.
\end{finding}

\part{Partition Analysis and the M4 Retraction}

\chapter{The M4 partition is a mid-training shape, not a stable feature}
\label{ch:m4-retraction}

\section{Original claim}

Session 12 day 1 analyzed \texttt{s12w1-lasso-pos-b2-s12-b4-20k} (M4) and
found a clean orthogonal partition: content in top-8 PCs (85\% variance,
zero position), position in PCs 8--19 (14\% variance, pos R$^2=0.29$). The
partition looked MDL-optimal---independent factors (content, position)
placed on orthogonal principal components with the larger factor on
higher-variance directions.

The initial reading: this is a stable architectural signature worth scaling
up.

\section{What W2-B showed}

\texttt{s12w2-lasso-pos-b2-s12-b4-50k} (W2-B) is the same architecture trained
to 50k steps. Its trajectory:

\begin{longtable}{rrrrr}
\toprule
step & EffDim & PC0-7 pos R$^2$ & PC8-19 pos R$^2$ & PC20-49 byte R$^2$ \\
\midrule
10k & 6.9 & 0.011 & 0.276 & 0.914 \\
20k & 5.8 & 0.071 & 0.294 & 0.560 \\
30k & 5.4 & 0.050 & 0.395 & 0.519 \\
40k & 5.5 & 0.040 & 0.386 & 0.534 \\
50k & 5.7 & 0.030 & 0.415 & 0.455 \\
\bottomrule
\end{longtable}

Three dynamics are visible:
\begin{enumerate}
  \item PC8-19 position R$^2$ grows monotonically from 0.276 at step 10k
  to 0.415 at step 50k. Position concentration in the mid-band intensifies
  with training.
  \item PC0-7 position R$^2$ leaks from near-zero to 0.07 at step 20k, then
  settles at 0.03--0.05. The ``orthogonal partition'' leaks slightly into
  the content band.
  \item PC20-49 byte-fingerprint R$^2$ decays from 0.914 to 0.455. The
  substrate trades byte-identity capacity in the tail for position
  representation.
\end{enumerate}

Comparing W2-B at step 50k to the single-band kin
\texttt{byte-lasso-pos-b2-s12-50k}:

\begin{longtable}{lrrr}
\toprule
metric & W2-B (b=4, 50k) & kin (b=1, 50k) & winner \\
\midrule
bpb (training log)  & 2.087 & 2.022 & kin ($-0.065$) \\
PC8-19 pos R$^2$    & 0.415 & 0.268 & W2-B (concentrated) \\
cum top-100 pos R$^2$ & 0.631 & 0.677 & kin (distributed) \\
EffDim               & 5.7  & 8.8   & kin \\
two-band flag        & NO/partial   & YES & kin \\
\bottomrule
\end{longtable}

\begin{retraction}[The M4 step-20k partition is a mid-training snapshot, not
a stable architecture]
The ``MDL-optimal two-band partition'' claim from session 12 day 1 described
a specific step-20k shape. With extended training the partition loses its
clean property: position leaks into the content band, content-fingerprint
capacity in deep PCs decays, and bpb underperforms the single-band kin. The
partition persists as a statistical feature but loses its clean orthogonal
structure. The single-band lasso-pos-b2-s12 family preserves a cleaner
partition at matched training.
\end{retraction}

The bpb gap is the operational decision: b=4 hits 2.087 at 50k, b=1 hits
2.022. Extending b=4 training further is unlikely to close the gap.

\section{Cross-check: cr-s12-pos-b2 at 50k shows the same pattern}

To verify this is not a b=4-specific anomaly, we ran the full step
trajectory of \texttt{s12w2-cr-s12-pos-b2-50k} (same complex-rotation family
at b=2):

\begin{longtable}{rrrrrr}
\toprule
step & k=8 & k=20 & k=50 & k=100 & PC8-19 \\
\midrule
10k & 0.000 & 0.057 & 0.285 & 0.402 & 0.057 \\
20k & 0.005 & 0.186 & 0.414 & 0.480 & 0.179 \\
30k & 0.004 & 0.251 & 0.473 & 0.512 & 0.245 \\
40k & 0.012 & 0.151 & 0.479 & 0.525 & 0.139 \\
50k & 0.014 & 0.122 & 0.450 & 0.481 & 0.108 \\
\bottomrule
\end{longtable}

PC8-19 position R$^2$ peaks at step 30k and decays afterward; cumulative
top-100 position R$^2$ peaks at step 40k and regresses slightly by 50k. Same
peak-and-regress pattern. The pos-b2 family's partition is
training-phase-dependent, not an architectural fixed point.

\chapter{The partition-score refinement}

\section{Binary flag failure mode}

Session 12 day 1's \texttt{two\_band\_partition} flag used the rule
\texttt{top8\_pos < 0.01 AND max\_tail\_pos > 0.05}. W2-B at step 50k has
\texttt{top8\_pos = 0.030} (above threshold) and \texttt{PC8-19 = 0.415}.
The flag flipped to NO despite a clearly partitioned substrate with a
$14\times$ tail/top ratio.

\begin{finding}[Binary partition-flag was too strict]
The session-12-day-1 \texttt{two\_band\_partition} flag missed partitions
that had any top-band leak, including well-established cases. Needed
replacement by a continuous score.
\end{finding}

\section{Continuous score and categorical verdict}

The replacement is a regularized ratio and a four-way verdict:

\begin{align*}
    \mathrm{score} &= \frac{\max_{\mathrm{tail}}\mathrm{pos\_r2}}%
                          {\max(\mathrm{top8\_pos\_r2}, 0.005)} \\
    \mathrm{verdict} &= \begin{cases}
        \text{absent}      & \max_{\mathrm{tail}} < 0.02 \\
        \text{leaky}       & \text{score} < 2 \\
        \text{partial}     & 2 \leq \text{score} < 5 \\
        \text{partitioned} & \text{score} \geq 5
    \end{cases}
\end{align*}

A regularization floor of 0.005 in the denominator prevents division by
zero when the top band is perfectly content-only; in that case the score
caps at $\max_{\mathrm{tail}}/0.005 \approx 200$ for a clean partition and
grows with tail strength.

Verdicts on the session-12 cohort:

\begin{longtable}{lrl}
\toprule
model & score & verdict \\
\midrule
s12w2-lasso-pos-b2-s12-b4 step 10k  &  24.5 & partitioned \\
s12w2-lasso-pos-b2-s12-b4 step 20k  &   4.1 & partial \\
s12w2-lasso-pos-b2-s12-b4 step 30k  &   7.9 & partitioned \\
s12w2-lasso-pos-b2-s12-b4 step 50k  &  13.9 & partitioned \\
s12w1-cr-s12-pos-b2 20k             &  59.9 & partitioned \\
s12w1-lasso-pos-b2-s8 20k           &  45.1 & partitioned \\
byte-lasso-pos-b2-s12-50k (kin)     &  53.7 & partitioned \\
byte-cr-hrrangle-s12-50k            &   0.0 & absent \\
cr-s8-scandead-20k                  &   0.0 & absent \\
\bottomrule
\end{longtable}

The back-compat \texttt{two\_band\_partition} boolean now triggers on
\texttt{partial OR partitioned}, which preserves its original intent
(``EffDim undercounts working dimensionality here'') while including the
leaky-but-real partition cases.

Four regression tests cover the verdict ranges: absent (content-only synthetic
substrate), partitioned (synthetic with orthogonal content/position bands),
leaky (synthetic with entangled content+position), and score-regularization
(edge case with exactly-zero top band). All pass. Commit \texttt{3a404f0}.

\chapter{The EffDim undercount, revisited}

Session 11 reported EffDim values that, we now know, substantially under-count
working dimensionality on any substrate that has two-band structure. The
partition analysis makes the mechanism specific:

\begin{itemize}
  \item EffDim is the entropy of the variance spectrum: $\exp(-\sum_i p_i \ln
  p_i)$ where $p_i$ is the $i$-th PC's fraction of variance. It describes
  the effective number of dimensions doing variance-weighted work.
  \item A two-band substrate places independent factors (content, position,
  byte-fingerprint) on disjoint PC ranges with different variance scales.
  Content occupies the high-variance head, position a mid-variance band,
  byte-fingerprint a low-variance tail with high selectivity but small overall
  contribution.
  \item EffDim sees the top band's eight modes and reports EffDim $\sim 8$.
  It doesn't count the position band because its variance share is small;
  it doesn't count the byte-fingerprint tail because their variance share is
  smaller still.
\end{itemize}

For the lasso-pos-b2 family, the working dimensionality is at least
$\sim$50 (head for content, mid-band for position, tail for
byte-fingerprint). EffDim reports values of 5--15. \emph{On any model
with simultaneously-nonzero content and position R$^2$, the EffDim
undercount is 3--10$\times$.}

\begin{retraction}[Session-11 quoted EffDim values on multi-factor substrates]
EffDim values reported in session 11 for the lasso-pos-b2 family, the
cr-hrrangle family, and the mut-persistent-s12 variants are
under-counts of working dimensionality by 3--10$\times$ whenever
ConR$^2 > 0.1$ and PosR$^2 > 0.01$ simultaneously. The replacement tool
is \texttt{heinrich profile-cb-pc-bands}, which reports per-band variance
and R$^2$ values separately and flags two-band partitions explicitly.
\end{retraction}

\part{Tools}

\chapter{New tools and plumbing added in session 12}
\label{ch:tools}

\section{\texttt{heinrich profile-cb-pc-bands}}

The primary new measurement tool. For a sequence-mode MRI, it reports
per-PC-band variance, per-band position R$^2$, per-band byte R$^2$, a
partition score, and a categorical verdict. See Chapter 5 for the score
definition. The tool handles substrate-primary and local-primary substrates
identically.

\begin{lstlisting}[caption={Example invocation},basicstyle=\ttfamily\small]
heinrich profile-cb-pc-bands --mris \
  /Volumes/Sharts/heinrich/session12/s12w1-lasso-pos-b2-s12-b4-20k.seq.mri \
  /Volumes/Sharts/heinrich/session11/byte-lasso-pos-b2-s12-50k.seq.mri

heinrich --json profile-cb-pc-bands --mris ... \
  | jq '.reports[] | {mri, partition_verdict, partition_score}'
\end{lstlisting}

\section{\texttt{heinrich mri-recapture}}

Bulk-recapture of stale MRIs. Reads \texttt{metadata.provenance.model\_path}
from each MRI's metadata JSON; falls back to naming convention if absent
(handles step-checkpoint patterns
\texttt{\{run\}-step\{N\}k.seq.mri $\leftarrow$ \{run\}-\{M\}k\_step\{N000\}.checkpoint.pt}).
Auto-corrects byte-level models whose recorded val\_data points to a
non-byte shard.

Dry-run by default; \texttt{--execute} triggers the actual captures.
Operational pass on \texttt{session11} took 94 stale MRIs to
\texttt{fix\_level=2} in one pass; 3 MRIs with idiosyncratic names were
skipped with clear reasons.

\section{MRI provenance and fingerprint}

Every MRI captured during session 12 records:

\begin{itemize}
  \item \texttt{metadata.provenance.model\_path}: absolute path to the
  source checkpoint.
  \item \texttt{metadata.provenance.result\_json}: path to the chronohorn
  result JSON (empty for non-decepticon backends).
  \item \texttt{metadata.provenance.tokenizer\_path}.
  \item \texttt{metadata.provenance.tool\_fingerprint}: a small dict with
  \texttt{heinrich\_sha} and \texttt{decepticons\_sha}---short git SHAs of
  both repositories at capture time.
\end{itemize}

Commits: \texttt{057707e} (provenance plumbing, fingerprint, raw-mode shape
fix, \texttt{mri-recapture}, additivity extension).

\section{Non-adaptive forward-path collapse}

The session-11 P2 deferred item. The non-adaptive branch of
\texttt{CausalBankInference.forward\_captured} reimplemented the model's
forward inline. Now collapsed: runs \texttt{self.\_model(x)} once and reads
stashed intermediates. Model-side stashes were added to
\texttt{\_linear\_logits} (non-adaptive branch) and \texttt{\_local\_logits}:

\begin{itemize}
  \item \texttt{\_last\_states\_nonadaptive} (after transforms + hash memory)
  \item \texttt{\_last\_x\_embed\_nonadaptive}
  \item \texttt{\_last\_band\_logits\_list} (if multi-band)
  \item \texttt{\_last\_local\_logits}
\end{itemize}

New regression fixture
\texttt{test\_forward\_captured\_matches\_training\_forward\_nonadaptive}
asserts max$|\Delta|$ = 0 between \texttt{model(x)} and
\texttt{forward\_captured(x).logits} on a non-adaptive config with local
path enabled. Verified bit-identical on three substrate-family variants
(gated\_delta, lasso, standard with bands). Decepticons commit
\texttt{a3fb77e}.

\section{Raw-mode substrate-shape bug}

\texttt{\_capture\_mri\_causal\_bank} pre-allocated \texttt{substrate\_states}
at \texttt{cfg.n\_modes}, but adaptive-substrate models return
\texttt{\_last\_features = concat(substrate\_modes, x\_embed)} whose trailing
axis is \texttt{n\_modes + embed\_dim}. Raw-mode captures of these models
errored. Fixed by probing shape from the first forward call and allocating
actual shape; metadata \texttt{hidden\_size} now reflects the captured width
rather than \texttt{n\_modes}. This matches how the sequence-mode path was
already written.

\section{Extended \texttt{profile-cb-additivity}}

The bpb-only additivity tool now optionally runs on EffDim / PosR$^2$ /
ConR$^2$ / \texttt{active\_frac} with per-metric noise floors (defaulted
from session-11 seed spreads, overridable via CLI flags). Reveals cases
where a mutation combination is bpb-additive but geometry-non-additive, or
vice versa. Default behavior (bpb only) unchanged for back-compat.

\part{Negative Results}

\chapter{What doesn't work for context extension}
\label{ch:negative}

\section{Extending EMA half-life}

Tested range: half-life from 2 to 512. Knee moves from 8 to 16 bytes;
saturation bpb drops 0.02. See Chapter 2.

\section{Adding position signal to scan-dead architectures}

\texttt{s12w2-scanablate-possig-s8-20k} adds a position-signal input
(concatenation of $\log(1+t)$ to gate inputs) to the scan-ablation recipe.
Expected: extended effective context via position-aware gating.

Measured:
\begin{itemize}
  \item Effective context: $\sim 8$ bytes (same as scan-ablation).
  \item Partition verdict: absent (no position structure in the substrate).
  \item bpb at 20k: 1.849 (vs scan-ablation 50k: 1.806).
\end{itemize}

\begin{finding}[Position signal without a working scan does not extend
context]
Position signal input needs an integrating mechanism to turn into position
structure in the substrate. With half-life 2, there is no such mechanism;
the position input is absorbed as a per-token modulation only. bpb improves
slightly because the pos input provides a fast per-token signal, but
effective context stays at 8 bytes.
\end{finding}

\section{Hash memory}

See Chapter 2. \texttt{mut-hashmem-fixed-20k} shows identical knee (16 bytes)
and similar saturation bpb (2.09) to \texttt{mut-control-mlp-20k} (no hash,
saturation 2.08). Hash memory is decorative for context extension.

\section{Local conv width (for local-primary family)}

Not directly tested (requires retraining; inference-time conv width change
is not supported). However, local-primary models at
\texttt{local\_window=8} show a 16-byte effective context; bumping to
\texttt{local\_window=32} would likely push the knee to 32--40 bytes and
drop bpb by $\sim 0.1$. Saturation ceiling would still be bounded by the
local-primary family at roughly 1.3--1.5 bpb at $k=32$.

\section{Feedforward conversion}

Tested: zero the substrate's contribution to readout features, measure
bpb. Result: bpb jumps from 1.80 to 6.65. The substrate is doing real
context work; scan-ablation is not a feedforward model in disguise. See
Chapter 3.

\section{Mode truncation}

Tested: zero modes beyond rank-$K$ at inference. Result: free $2\times$
reduction (2048 $\to$ 1024 modes at identical bpb); sharp degradation
below $K=256$. Low-rank $\omega$ (rank 28) does not imply low-rank
substrate. Useful efficiency lever but not a $10\times$ jump. See Chapter 3.

\part{The One Remaining Experiment}

\chapter{Substrate-primary gated-delta}

\section{What hasn't been tested}

Every gated-delta model in the library has \texttt{local\_scale=0.25}
(local-primary). Every substrate-primary model has
\texttt{substrate\_mode=frozen} (adaptive HRR basis). The union has never
been run:

\begin{lstlisting}[caption={The untested recipe},basicstyle=\ttfamily\small]
--substrate-mode gated_delta
--adaptive-substrate
--local-scale-override 0.0
--linear-half-life-max 512
--hrr-omega-init
--linear-readout-kind routed_sqrelu_experts
--linear-readout-num-experts 2
--scale 8.0
--steps 50000
\end{lstlisting}

This is scan-ablation's recipe with \texttt{substrate\_mode} swapped from
frozen to gated\_delta. One flag change. It tests whether the Mamba-class
selective gating mechanism (per-position input-dependent retain/write/erase
gates) breaks the substrate-primary context ceiling that the frozen-HRR
substrate hits at 16 bytes.

\section{What the outcome decides}

The decision tree:

\begin{itemize}
  \item \textbf{If knee $\leq 16$ bytes and saturation $\geq 1.78$ bpb}:
  gated-delta selective gating does not escape the substrate-primary
  ceiling. The causal-bank architecture family is context-ceilinged at
  $\sim 16$ bytes regardless of mechanism. Pursuit shifts to hybrid
  architectures---specifically, windowed attention with $k \leq 64$ (still
  $O(n\cdot k)$ but not $O(n^2)$) or a substantially different integrative
  mechanism.
  \item \textbf{If knee $\geq 32$ bytes or saturation below 1.5 bpb}:
  gated-delta selective gating escapes the ceiling. The causal-bank family
  has genuine long-context capacity. Scale it up; pursue scale-14
  architectures; the 1.0 bpb target is reachable within this family.
  \item \textbf{If something in between}: partial lift. Characterize the
  gain per param and decide whether the family is worth further investment
  vs hybrid-architecture exploration.
\end{itemize}

\section{Prior on the outcome}

Honest estimate: 50/50.

Arguments for escape: gated-delta has per-position input-dependent gates
that can selectively retain specific past events rather than blur them into
an EMA summary. Gradient routing through such gates is known
(Mamba-literature) to propagate long-range dependency.

Arguments against escape: decepticons' \texttt{\_gated\_delta\_states}
implementation uses sigmoid-gated retain/write/erase with a chunked parallel
scan. Already tested in local-primary configuration; that configuration
has the same 16-byte ceiling as frozen-HRR substrate. The ceiling may be
intrinsic to the gradient-routing story regardless of the specific gating
parameterization.

A reasonable bet: if chronohorn runs this pilot at 50k steps on s8, the
outcome is knowable in roughly 30 minutes of A4000 compute. It is the only
remaining scoped experiment that would unambiguously decide the future of
the causal-bank architecture family for long-context byte modeling.

\part{Retrospective}

\chapter{Retractions}

\begin{retraction}[``M4's step-20k partition is the architectural target'']
See Chapter 4. The M4 partition is a mid-training shape, not a stable
feature. b=4 factorization underperforms b=1 at matched training.
\end{retraction}

\begin{retraction}[``Longer EMA half-life translates to longer effective context'']
Tested: half-life 2 gives 8-byte knee; half-life 512 gives 16-byte knee.
256$\times$ more theoretical memory buys 2$\times$ actual context. See
Chapter 2.
\end{retraction}

\begin{retraction}[``Speed arm is one architecture'']
Speed arm is two architectures with structurally different prediction
machinery. See Chapter 1.
\end{retraction}

\begin{retraction}[``Hash memory's session-11 training fix may extend context'']
Hash memory, post-fix, has the same 16-byte knee as the hash-less control.
Retrieved context is too noisy to extend the effective window.
\end{retraction}

\begin{retraction}[``Scan-ablation might be feedforward and $\omega$-dominated'']
Zeroing substrate drives bpb 1.80 $\to$ 6.65. The substrate is doing real
context work. See Chapter 3.
\end{retraction}

\begin{retraction}[``$\omega$ low rank implies substrate low rank'']
$\omega$'s rank-28 projection is input-side. The substrate itself uses
$\sim$1024 modes meaningfully. Truncating beyond $K=256$ degrades bpb.
See Chapter 3.
\end{retraction}

\chapter{What's still unknown}

\begin{itemize}
  \item Whether gated-delta substrate-primary breaks the 16-byte ceiling
  (Chapter 8). The single experiment that remains.
  \item Whether sliding-window attention at $k \in \{32, 64\}$ would give
  a better bpb/FLOP tradeoff than the above. Not piloted; estimated by
  scaling laws.
  \item What the byte-level transformer baseline at matched params actually
  is on FineWeb-10B. Literature prior 1.1--1.3 bpb at 36M params; not
  directly measured in this library.
  \item Whether the substrate-primary bpb ceiling at 1.78 is tight against
  the English 8-gram entropy bound, or whether there's still headroom in
  per-token encoding capacity. Can be tested by swapping the routed-experts
  readout for a richer parameterization at matched FLOPs.
\end{itemize}

\chapter{The plumbing register}

Fixes and additions landed in session 12:

\begin{description}
  \item[Heinrich \texttt{057707e}] Raw-mode substrate-shape bug, MRI provenance
  (model\_path / result\_json / tokenizer\_path), tool fingerprint
  (heinrich + decepticons SHAs), \texttt{heinrich mri-recapture} CLI,
  extended \texttt{profile-cb-additivity} (bpb + EffDim + PosR$^2$ + ConR$^2$
  + active\_frac metrics).
  \item[Heinrich \texttt{e160516}] \texttt{heinrich profile-cb-pc-bands} CLI
  with per-band variance, position R$^2$, byte R$^2$ reporting.
  \item[Heinrich \texttt{3a404f0}] Partition-score refactor: binary
  \texttt{two\_band\_partition} replaced with continuous
  \texttt{partition\_score} and categorical \texttt{partition\_verdict};
  four new regression tests.
  \item[Decepticons \texttt{a3fb77e}] Non-adaptive forward-path collapse
  (session-11 P2 deferred). Model-side stashes for substrate /
  x\_embed / band\_logits / local\_logits in eval mode; loader reads them.
  Regression fixture asserts bit-identical.
\end{description}

\chapter{Recommendations for chronohorn}

\begin{enumerate}
  \item \textbf{Retire the ``speed arm'' nomenclature} until architecture-family
  tagging is in place. Every cross-model comparison needs to distinguish
  substrate-primary from local-primary.

  \item \textbf{Add \texttt{architectural\_family} tag at training time}.
  Set to \texttt{substrate-primary} if \texttt{local\_scale == 0}, else
  \texttt{local-primary}. Persist in both result JSON and checkpoint
  metadata.

  \item \textbf{Run the substrate-primary gated-delta pilot} (Chapter 8).
  One flag change from scan-ablation; $\sim$30 min on A4000; resolves the
  largest remaining architectural question.

  \item \textbf{Retrain scan-ablation-s8 at
  \texttt{--linear-modes=1024}} (Chapter 3). Free 2$\times$ param reduction
  at identical bpb; 92k $\to$ $\sim$150--180k tok/s. No architectural
  change.

  \item \textbf{Enforce \texttt{test\_bytes\_per\_token} non-null in result
  JSON ingestion}. Prevents frontier-class contamination (sp-tokenized models
  appearing on the byte-level frontier table under hardcoded conversion).

  \item \textbf{Add NaN tripwire at checkpoint save}. Prevents
  NaN-diverged runs from polluting the library. See
  \texttt{byte-hrr-frozen-s12-50k.checkpoint.pt} (session 11) for an
  example---21 of 30 tensors NaN-saturated, captured cleanly but the
  model is cooked.

  \item \textbf{Add chronohorn commit SHA to result JSON} at train time.
  Heinrich records its own + decepticons' SHA on every MRI; the training
  pipeline's SHA should propagate too for full reproducibility.

  \item \textbf{Drop the local-window experiment} (bumping local-conv window
  on local-primary) until it becomes the critical path. The local-primary
  family caps at 1.3--1.5 bpb and is bounded below the transformer target
  by construction. Useful for deployment efficiency, not the 1.0 bpb target.
\end{enumerate}

\chapter{What this book is not}

A tutorial on Heinrich, chronohorn, or decepticons. A survey of the
causal-bank architecture family. An introduction to the EMA substrate, the
HRR basis, or the gated-delta mechanism. A statement about whether the
1.0 bpb byte-level target is reachable in general.

It is a record of one session's forensic work. The central finding is an
architectural distinction that had been hiding in plain sight; the
secondary finding is a structural context ceiling that kills a family of
proposed mutations; the tertiary finding is a methodological clarification
(continuous partition score, EffDim undercount) that rescues analysis
tooling from false negatives.

The big thread across session 11 and session 12: measurement apparatus
failure modes are most dangerous when they agree with the interpretation
you'd like to report. A 16-byte ceiling that looks architectural could be a
measurement artifact; a measurement artifact that looks architectural will
ship as a published finding. Both happened at different scales in this
session. The discipline is to keep asking ``what would prove this wrong''
until the answer is ``one pilot at 30 minutes of compute, and we'll know.''

We leave that pilot for chronohorn.

\end{document}
