\documentclass[11pt,a4paper]{article}
\usepackage{fontspec}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{float}
\usepackage{longtable}
\geometry{margin=1in}

% --- The provenance apparatus, inherited from The Frozen Frame: every number
% states the artifact it was measured from and the code that produced it. In a
% paper whose subject is a number we believed too easily, this is not decoration. ---
\definecolor{prov}{RGB}{95,95,95}
\newcommand{\provenance}[1]{\par\vspace{2pt}{\footnotesize\ttfamily\raggedright\color{prov}#1\par}}
\newcommand{\heinrich}{\textsc{Heinrich}}
\newcommand{\chronohorn}{\textsc{Chronohorn}}
\newcommand{\decepticons}{\textsc{Decepticons}}
\newcommand{\bpb}{\,bpb}

\title{Counting Beats the Organ\\[0.5em]
\large A Negative Result on Retrieval-Augmented Causal Banks,\\
and the Baseline We Never Ran}

\author{Built by Claude instances, steered by a human, measured by machines\\
Systems: \decepticons{} (kernel) $\cdot$ \chronohorn{} (runtime) $\cdot$ \heinrich{} (forensics)\\[0.5em]
\small Every number below was co-witnessed by two independent instruments\\
and none of them was worth having.}

\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
We set out to buy bits. A frozen-substrate causal bank---a recurrent body whose
kernel is fixed at initialization and whose readout is trained---was to be given a
memory organ: a $k$-nearest-neighbour datastore over its own hidden states, mixed
into its output distribution. Over five store sizes from 8M to 150M bytes we
measured the organ's value, co-witnessing every rung to the fourth decimal across
two independent implementations, and it behaved: the lift grew monotonically, the
optimal mixing weight rose with store size, an oracle bound priced the ceiling, and
a four-corner control experiment cleanly separated what retrieval was buying
(calibration repair) from what it was not (cross-domain knowledge). Then we ran, at
the operator's suggestion, the control we should have run first. \textbf{An
order-7 byte $n$-gram---a count table, ninety seconds of CPU---dominates the organ at
every store size, does not saturate, and, alone, out-predicts the trained model
itself} (2.0197 vs.\ 2.0367\bpb). Our best mix, 150M keys and 38\,GB of RAM at
1.9530\bpb, is beaten by a count table holding a third of the bytes (1.8681\bpb).
Independently and from the opposite end, a weight-level measurement finds that the
trained readout places \emph{less} energy in the directions that decode the lagged
past than a random matrix would ($0.40$--$0.43\times$ chance): training does not
merely fail to use long context---it turns away from it. The two results are the
same fact seen from two sides, and the fact is that the body was not using its
long context, so an organ that supplies more of it could not have helped. We report
the negative result, the one claim that survives it (a byte unigram---256
numbers---captures 93.6\% of what a 64-million-key archive buys on foreign text),
and the methodological failure that produced a month of honest, precise,
irrelevant measurement: every pre-registration we wrote carried a literature line
and a kill line, and none carried a \emph{baseline} line.
\end{abstract}

\section{What This Document Is}

This is a negative result, reported in full, by the two agents who produced it and
the human who ended it with one question.

Three repositories are involved and it matters which does what. \decepticons{}
supplies the kernel: the causal-bank model, its frozen linear substrate, and the
retrieval primitives. \chronohorn{} supplies the runtime: training, evaluation
harnesses, and the datastore experiments. \heinrich{} is the forensic instrument:
it measures what a model \emph{computes}---residual geometry, state information,
weight structure---rather than what it outputs, and it was used here both to
co-witness \chronohorn{}'s claims on an independent implementation and to make
measurements no loss curve can make.

The campaign had a target: 1.0 bits per byte on \texttt{enwik8}-family text with a
small byte-level model. It had a strategy: since the substrate is frozen and cannot
learn to store the past, \emph{give} it the past, as an external archive of its own
states. It had a protocol we were proud of---pre-registration before every run, a
declared kill line, two independent witnesses agreeing to the fourth decimal before
any number entered the record. Section \ref{sec:campaign} reports what that
protocol produced. Sections \ref{sec:ngram} and \ref{sec:evacuation} report the two
measurements that destroyed it. Section \ref{sec:anatomy} is the part we think is
actually worth publishing.

\section{The Bet}

The causal bank is a recurrent architecture whose state is a bank of exponential
moving averages of a projected byte embedding, with decay rates fixed at
initialization across a spectrum of half-lives. The substrate is \emph{frozen}: it
is never trained. Only a readout---an MLP reading the concatenation of the state and
a local embedding---learns. The architectural bet is that a rich enough fixed
temporal basis, plus a trained readout, can substitute for a learned recurrence,
and that this buys enormous efficiency.

The bet's obvious weakness is memory. A frozen EMA bank cannot choose what to keep.
Long-range structure that a transformer holds verbatim in its KV cache, or that an
LSTM learns to gate, must here be recoverable from a fixed linear mixture of the
past---or not at all. This is the gap the retrieval organ was meant to close: if
the body cannot hold the past, let it \emph{look the past up}.

Concretely: run the body over a large corpus, store each state vector alongside the
byte that followed it, and at inference retrieve the $k$ nearest stored states to
the current one, vote over their successor bytes, and mix that vote into the body's
output distribution with weight $\lambda$. This is $k$NN-LM, transplanted from
transformers onto a causal bank, and the ladder that follows measures what it buys.

\section{The Campaign, As It Looked From Inside}
\label{sec:campaign}

\subsection{The ladder}

The base model is \texttt{fo-noadapt-enwik90-50k}: a byte-level causal bank trained
on the first 90M bytes of \texttt{enwik9}, evaluated on held-out
\texttt{enwik8}-family windows at \textbf{2.0367\bpb}. Every rung below stores
states from the same corpus and queries held-out windows; every rung was run twice,
once in \chronohorn{}'s harness and once in \heinrich{}'s independent
implementation with its own cross-entropy accounting, and the two agreed to the
fourth decimal at all five rungs.

\begin{table}[H]
\centering
\begin{tabular}{lrrr}
\toprule
store & $\Delta$ vs.\ base & $\lambda$ & mix \\
\midrule
8M   & $-0.0144$ & 0.05 & 2.0223 \\
16M  & $-0.0187$ & 0.05 & 2.0180 \\
32M  & $-0.0367$ & 0.10 & 2.0000 \\
64M  & $-0.0488$ & 0.10 & 1.9880 \\
94M  & $-0.0686$ & 0.20 & 1.9682 \\
150M & $-0.0837$ & 0.50 & 1.9530 \\
\bottomrule
\end{tabular}
\caption{The kNN ladder. Monotone, double-witnessed, and beside the point.}
\end{table}
\provenance{goldens: docs/data/knn-ladder-*.txt \quad{} base guard: 2.0367 exact or abort}

It behaved exactly as a working mechanism should. The lift grew with store size.
The optimal mixing weight $\lambda$ rose with it---roughly doubling per two store
doublings---which we read as a demand curve: the bigger the archive, the more the
model wants to listen to it. Two independently constructed calibration grids, with
different geometries, never disagreed about the shape of that curve.

\subsection{The oracle purse, and a case for a port}

For each rung we computed a per-position oracle over the two endpoints
$\{$base, pure $k$NN vote$\}$: an upper bound on what \emph{any} gating function
$\lambda(x)$ could collect. The purse grew with the store (0.2610\bpb at 8M,
0.3788 at 94M, 0.4377 at 150M) while the scalar mixture's take of it plateaued
around 19\%. We called this the \emph{widening gap} and it became the strongest
quantitative argument for the next experiment---an attention port that would read
the retrieved neighbours directly instead of averaging them into a vote. The
argument had the pleasing property of not depending on the oracle being achievable:
it rested only on the mixture demonstrably failing to track a growing purse.

It was an artifact. See \S\ref{sec:ngram}.

\subsection{The relevance square, and the surface table}

The one part of the campaign that still stands as \emph{method} is the
decomposition. A raw retrieval delta conflates two different purchases: an archive
of \emph{any} text repairs a badly calibrated base (\emph{smoothing}), and an
archive of \emph{relevant} text supplies content the base lacks (\emph{knowledge}).
Running a store of irrelevant text as a control isolates the first, and the
difference gives the second.

Across four held-out domains, with an \texttt{enwik8} store as the irrelevant
control:

\begin{table}[H]
\centering
\begin{tabular}{lrrrr}
\toprule
fold & base & total & smoothing & knowledge \\
\midrule
code   & 4.9036 & $-2.9080$ & $-1.1837$ & $\mathbf{-1.7243}$ \\
math   & --     & $-0.7666$ & $-0.3233$ & $\mathbf{-0.4433}$ \\
wikiml & --     & $-0.4566$ & $-0.5740$ & $+0.1174$ \\
world  & 4.8797 & $-0.4204$ & $-0.6245$ & $+0.2041$ \\
\bottomrule
\end{tabular}
\caption{The surface table. On foreign prose, \emph{relevance subtracts}: the
irrelevant store beats the relevant one. On templated text (code, math) the
knowledge column is real and large.}
\end{table}
\provenance{goldens: docs/data/code-fold-cowitness.txt, knn-square-*.txt \quad{} code fold + control co-witnessed at the 4th decimal}

The sign of the knowledge column separates two regimes cleanly, and a cheap
label-free discriminator---is the archive alone a better predictor than the base?
---predicted that sign correctly in all four folds without needing the control at
all. The policy sentence we drew from it was: \emph{build stores for templated
domains; calibrate for foreign ones.}

Hold onto the word \emph{calibrate}. It is about to become the whole story.

\section{Refutation I: A Byte Count Table Dominates the Organ}
\label{sec:ngram}

The operator asked why we had never used $n$-grams.

The question stings because the answer is not ``we lacked them.'' \decepticons{}
ships \texttt{OnlineCausalMemory}, a causal $n$-gram accumulator. The polyhash
module ships \texttt{NgramTable}. We had count tables in both repositories the
entire time and never pointed one at the organ.

So we built the control properly: an interpolated byte $n$-gram over orders $0..K$
with Witten-Bell-style backoff and 64-bit rolling context hashes, trained on
\emph{the same store bytes} a $k$NN rung uses, scored on \emph{the same query
windows}, mixed into \emph{the same base logits}. Two numbers decide it: which is
the better standalone predictor, and which buys more when mixed.

\subsection{Standalone}

\begin{table}[H]
\centering
\begin{tabular}{lr}
\toprule
predictor & bpb \\
\midrule
best byte $n$-gram, \textbf{8M} store (order 7, $\alpha{=}64$) & \textbf{2.3128} \\
state-$k$NN organ alone, 8M store & 2.9743 \\
state-$k$NN organ alone, \textbf{150M} store & 2.3892 \\
\midrule
the trained base itself & 2.0367 \\
$n$-gram alone, 94M store & $\mathbf{2.0197}$ \\
\bottomrule
\end{tabular}
\caption{An order-7 count table on 8M bytes beats the retrieval organ on 150M
bytes. At matched store size it is not close. And at 94M, the count table
out-predicts \emph{the trained model}.}
\end{table}
\provenance{chronohorn: python/chronohorn/eval/ngram\_control.py \quad{} ruling b2815ef5}

\subsection{Mixed---the decisive test}

Standalone strength is not the claim the ladder made. The ladder measured what the
organ buys \emph{mixed into a trained base}, and complementarity is not strength:
the base is a neural LM trained on this very corpus and may already contain every
$n$-gram statistic there is, in which case mixing them back in would buy nothing.
That was the organ's last defence, and it fails.

\begin{table}[H]
\centering
\begin{tabular}{lrrr}
\toprule
store & organ $\Delta$ & $n$-gram $\Delta$ & ratio \\
\midrule
8M  & $-0.0144$ & $\mathbf{-0.0515}$ & 3.6$\times$ \\
94M & $-0.0686$ & $\mathbf{-0.1686}$ & 2.5$\times$ \\
\bottomrule
\end{tabular}
\caption{Same body, same store bytes, same canonical eval windows (the base must
reproduce 2.0367 exactly or the run aborts). The $n$-gram does not saturate: it
scales alongside the organ from a far higher base, so the organ never catches up.}
\end{table}
\provenance{chronohorn commit bd68f85 \quad{} ruling abff0759 (reset)}

\textbf{base $+$ $n$-gram at 94M $=$ 1.8681\bpb beats base $+$ organ at 150M $=$
1.9530\bpb}---our best-ever mix, beaten by a count table with a third of the bytes,
in ninety seconds of CPU and 5.3\,GB, against hours of GPU and 38\,GB.

\subsection{What this retracts}

The \emph{widening gap}---our strongest argument for the attention port---dies
here, and it dies in the most embarrassing way available: the scalar mixture
collects the $n$-gram at \textbf{41.4\%} of its purse where it collects the organ
at \textbf{18.1\%}. The mixture was never structurally blind. It was being handed a
weak, badly calibrated predictor. The ``unrealized purse'' we were pointing at was
\emph{the organ's own incompetence}, wearing the costume of a mechanism.

Dead, therefore: the five-rung ladder as a value claim; the product-quantization
tier that would have built 900M keys; the attention port's premise as written. The
ladder measured the \emph{cost} of retrieval, not its value.

\subsection{The near-miss that should frighten us most}

The first pass of this control used $\alpha = 1.0$, untuned, and reported that
\emph{the organ beats the $n$-gram by 0.345\bpb}. Pushing $\alpha$ until the curve
turned inverted the verdict by \textbf{1.03\bpb}.

We came within one lazy hyperparameter of publishing the opposite conclusion, with
four decimals and two witnesses behind it. \textbf{An untuned control is not a
control; it is a prop.} The rule we extract in \S\ref{sec:anatomy} is stronger than
``run a baseline,'' and this is why.

\section{Refutation II: The Readout Evacuates the Past}
\label{sec:evacuation}

The second refutation was not looking for the first. It came from \heinrich{}, from
a weight measurement, and it arrived on the same day.

\subsection{The question, and the null}

The frozen substrate holds the past in a fixed linear mixture. \emph{Which
directions of the state actually decode the lagged past?} We answered it by
frequency-weighted class-mean SVD on a sequence-mode capture of the body's own
frozen initialization---for each lag $\ell$, the directions along which the state
separates by the byte that occurred $\ell$ steps ago. For the experimental body
(512 modes) this yields a rank-7 subspace: 5 directions at lag 64, 2 at lag 512,
merged and orthonormalized.

Then, the crucial part, and the part we nearly got wrong. \emph{How much energy
would a readout put in those 7 directions by accident?} An isotropic (Xavier) init
already places a fraction $r/D$ of its weight energy in \emph{any} fixed rank-$r$
subspace. Here $7/512 = \mathbf{0.0137}$. The measured value on the real untrained
body is $0.0131$. \textbf{An untrained readout is already as ``aligned'' with the
retrospective subspace as chance allows}, and any alignment claim not quoted
against this floor is unreadable. (We initially carried a floor of $0.109$ from a
64-mode toy---a number correct in its own context and meaningless in this one. The
error is the same class as everything else in this paper.)

\subsection{The measurement}

Four training arms, identical but for how the readout's state-facing block is
seeded: naive (Xavier), zeroed, and two \emph{projected} arms that rotate a
prescribed fraction of the block's energy into the retrospective subspace while
conserving its Frobenius norm exactly---$1/3$ (a $24.4\times$ chance
concentration) and $0.05$ ($3.7\times$).

\begin{table}[H]
\centering
\begin{tabular}{lrrrr}
\toprule
checkpoint & energy in subspace & $\times$ chance & $\|$block$\|$ & eff.\ rank \\
\midrule
init (pre-hook)   & 0.0131 & 0.96 & 17.73 & 170.60 \\
\midrule
A naive @1k       & 0.0087 & 0.64 & 31.07 & 40.27 \\
A naive @5k       & 0.0056 & \textbf{0.41} & 48.36 & 33.44 \\
B zeroed @1k      & 0.0104 & 0.76 & 23.50 & 18.91 \\
B zeroed @5k      & 0.0059 & \textbf{0.43} & 41.15 & 25.94 \\
C reserve @1k     & 0.0727 & 5.32 & 31.03 & 40.33 \\
C reserve @5k     & 0.0138 & \textbf{1.01} & 48.50 & 36.04 \\
C2 mid @1k        & 0.0119 & 0.87 & 31.03 & 40.13 \\
C2 mid @5k        & 0.0054 & \textbf{0.40} & 47.35 & 35.50 \\
\bottomrule
\end{tabular}
\caption{The readout's state-facing block, projected into the rank-7 retrospective
subspace. Chance is $1.00\times$.}
\end{table}
\provenance{golden: docs/data/e4-init-geometry-survival.txt \quad{} instrument: heinrich profile-cb-subspace-energy \quad{} ruling 75906b12}

Two things are visible and only one of them was predicted.

\textbf{The imposed geometry dissolves.} Arm C begins at $24.4\times$ chance and
decays to $5.32\times$ by step 1000 and $1.01\times$ by step 5000. Whatever we
seed, training returns the readout to chance. The final losses of all four arms
(2.4107, 2.4103, 2.4061, 2.4107) lie inside the $\sim$0.004\bpb init-noise floor:
\emph{the init geometry does not survive training.} That much we expected to read
off the loss curves; here it is read off the weights.

\textbf{But the trained readout does not stop at chance---it goes below it.} Arms
A, B and C2 all finish at $0.40$--$0.43\times$ chance. \emph{The trained readout
places less energy in the directions that decode the lagged past than a random
matrix would.} It does not merely fail to use long context. It \emph{turns away
from it}, systematically, and it spends its training steps doing so.

This is the same arrow the forensic program had already found in the state's
information content---that training moves the body from an archive with its head in
the past toward attention on the present edge---now visible in a second, independent
surface: the readout's own weights.

\subsection{What we withdrew}

We had a story for why the aggressive projected arm (C) was slowest early:
over-concentration, i.e.\ rank collapse in the readout. It is \emph{not supported}.
At step 1000 arm C's effective rank (40.33) equals naive's (40.27). And our
evidence cannot see the regime where the effect lived---C's early penalty is at
steps 50--200, the earliest checkpoint is 1000, and by then the loss curves have
converged. The mechanism of C's early penalty is \textbf{unknown}. We record it as
unknown, having proposed it loudly and having had a collaborator build an
experimental arm around it.

\section{Convergence: The Body Is Not Using Its Long Context}

Put the two refutations side by side.

\begin{itemize}[leftmargin=1.5em]
\item A 7-byte count table out-predicts the entire trained body (2.0197 vs.\ 2.0367).
\item The trained readout actively turns away from long-lag directions (0.41$\times$ chance).
\end{itemize}

They are the same fact from opposite ends. The first says: whatever the body earns
above the count table's reach, it does not earn from context longer than about
seven bytes. The second says: the body's readout has \emph{decided} not to look
there. The frozen substrate's long-timescale modes---the entire architectural
premise, the thing the retrieval organ was built to supplement---may be dead weight
the readout has learned to ignore.

This reframes the negative result. The organ did not fail because retrieval is
worthless, or because $k$NN is the wrong retrieval mechanism. \textbf{It failed
because it was supplying more of a resource the body had already refused.} No
better organ, no attention port, no product-quantized 900M-key store would have
changed that, and each was on our roadmap.

\section{What Survives}

\subsection{The unigram prices the smoothing column}

The single most useful number the campaign produced is the one that embarrassed it.
On a foreign fold (\texttt{enwik8} store, \texttt{world} queries, base 4.8797), we
scored the archive against the store corpus's \textbf{byte marginal}---256 numbers,
no keys, no search---on identical positions, identical base logits, with $\lambda$
tuned on the same calibration split:

\begin{table}[H]
\centering
\begin{tabular}{lrr}
\toprule
mixed with & $\Delta$ & mix \\
\midrule
the archive (64M keys) & $-0.6245$ & 4.2553 \\
a byte unigram (256 numbers) & $-0.5843$ & 4.2955 \\
\bottomrule
\end{tabular}
\caption{The unigram captures \textbf{93.6\%} of what a 64-million-key archive
buys. Co-witnessed on two instruments; the figure reproduces to the digit.}
\end{table}
\provenance{golden: docs/data/marginal-unigram-cowitness.txt \quad{} ruling ac7bc509}

This does not kill the smoothing/knowledge decomposition---it \emph{prices} it. What
we had been calling ``calibration repair'' was never retrieval at all: it is an
overconfident base being pulled toward byte frequencies, and \emph{anything} that
pulls it there works. An archive is an extraordinarily expensive way to compute a
unigram. The policy sentence survives with a cost column attached: for a smoothing
fold, do not build a store---ship 256 numbers.

\subsection{The decomposition, as method}

The square and the smoothing/knowledge split are claims about \emph{mixing an
archive into a base}, not about state-$k$NN. They survive the reset as a method,
and every number in them must be re-measured with a counting archive. The open
question it leaves is sharp and cheap: templated text is exactly what $n$-grams
eat, so \emph{does code's $-1.7243$ knowledge column survive against a count-table
archive?} If it does, relevance is real. If it does not, the square was about count
statistics all along.

\subsection{The instruments}

The forensic tools outlive the hypothesis they were built to test:
\texttt{profile-cb-subspace-energy} (does an init geometry survive training?),
\texttt{profile-cb-retro-subspace} (which state directions decode the lagged
past?), and a $k$NN-lift harness that now \emph{always} emits its own unigram
control. That last change is the paper's one line of code worth copying: the
refutation is no longer optional, so it cannot be forgotten.

\section{Anatomy of an Error}
\label{sec:anatomy}

This is the part we believe is worth publishing.

The campaign was not sloppy. It was, by every local measure, unusually careful:
pre-registration before each run, a declared kill line, two independent
implementations agreeing to the fourth decimal before anything entered the record,
a written protocol for hardware contention, and an explicit rule that no analysis
survives as a script---it becomes an instrument or it does not exist. We caught, in
flight, a corpus loader silently interleaving nulls (by hashing bytes against their
producing reader), a GPU throttled 8$\times$ while reporting 100\% utilization, and
an evaluation harness in which \emph{raising the calibration count silently drew a
disjoint test set}---caught only because a guard tied the new harness to a known
base value and refused to run when it drifted by 0.03\bpb.

And all of that precision was spent measuring a mechanism that a ninety-second
count table beats.

\subsection{The rule}

\begin{quote}
\textbf{No experiment without its cheapest competent baseline, tuned to its own
optimum---and the baseline is tuned by whoever stands to lose from it.}
\end{quote}

Every pre-registration we wrote had a literature line and a kill line. None had a
\emph{baseline} line. The kill line asks ``what result would make me abandon this
hypothesis?''---and we answered it honestly every time, which is exactly why it felt
like rigour. But a kill line polices the hypothesis \emph{against itself}. It never
asks the other question: \emph{what cheap thing might already do this, and would I
notice if it did?}

The second clause is not rhetoric. The untuned $\alpha{=}1.0$ pass had the $n$-gram
losing by 0.345\bpb; tuning it flipped the verdict by 1.03. A baseline tuned by its
proponent's opponent is the only kind you can trust, and in a two-agent system that
is a role you can actually assign.

\subsection{The error class, five times in one day}

Once named, the same failure was everywhere---always a number that meant one thing
in one regime, quoted as if it meant the same in another:

\begin{enumerate}[leftmargin=1.5em,itemsep=2pt]
\item An untuned $n$-gram ($\alpha{=}1.0$) reported as ``the organ wins.''
\item A chance floor measured on a 64-mode toy ($0.109$) carried into a 512-mode
body, where the true floor is $0.0137$---an $8\times$ error in the null of the
central alignment claim.
\item A purse-anatomy figure remembered from a letter and compared against a fresh
measurement, nearly producing a spurious ``whitespace collapse.''
\item Gate rules posting an 84--86\% purse share on a lopsided fold---which we
nearly cashed as evidence of learned routing, until we read the usage column: the
gates were firing 94--96\% of the time. They had degenerated to
\emph{always-retrieve}, which is near-optimal when retrieval alone beats the base.
No routing was being learned at all. \emph{Read the usage before crediting the
share.}
\item An API argument written from memory of its own signature rather than copied,
silently mis-passed.
\end{enumerate}

\subsection{The fourth decimal is not rigour}

It is the \emph{last} step of rigour, and we performed it first and hardest because
it is the one that feels like virtue. Two witnesses agreeing to four decimals is a
strong guarantee that a number is \emph{correct}. It is no guarantee whatsoever that
the number is \emph{about anything}. We had a protocol optimized end-to-end for
precision and containing no step whose job was relevance.

Rigour is asking what would make the entire enterprise pointless, and then spending
ninety seconds of CPU to find out.

\section{Threats to Validity, and What We Did Not Show}

We are reporting a negative result and the temptation to over-generalize it is
exactly symmetrical to the temptation that produced the positive one. So, plainly:

\begin{itemize}[leftmargin=1.5em,itemsep=3pt]
\item \textbf{One architecture, one scale, one corpus family.} All of this is a
byte-level causal bank with a frozen substrate, 0.4M--11M parameters, on
\texttt{enwik}-family text. A count table beating \emph{this} body says nothing
about count tables beating a transformer, or a larger bank. Everything above
$\sim$1.8\bpb on this corpus is commodity statistics---PPM's well---and our best mix
(1.9530) never left the well. A model that had left it might use its long context.
\item \textbf{The organ was \emph{our} organ.} Whitened state $k$NN with a softmax
vote is one retrieval design. A different key construction might not be dominated.
We showed that \emph{this} organ is dominated, and that the reason is upstream of
the organ---but the second half of that claim rests on the readout measurement,
which is a single body.
\item \textbf{E4 is one seed per arm.} The early-training inversions are
46--69$\times$ the noise floor, so they should hold, but they are not replicated.
The steps 50--200 regime, where the only unexplained effect lives, was never
captured.
\item \textbf{The context-length decomposition was never run.} Our central claim
---the body is not using its long context---rests on two convergent measurements,
not on a direct one. The direct measurement is stated below and it is cheap. We did
not run it, and a reader is entitled to hold the claim at exactly the confidence two
indirect witnesses earn.
\end{itemize}

\section{Conclusion, and the One Question Left}

We built a retrieval organ for a frozen-substrate causal bank, measured it across
six store sizes with two independent instruments to the fourth decimal, priced its
ceiling with an oracle, decomposed its value into smoothing and knowledge with a
four-corner control, and designed an attention port to collect the residual. A byte
count table beats it at every size, does not saturate, and out-predicts the model
itself. The model, independently, turns out to place less weight on the directions
that decode its own distant past than a random matrix would.

The organ was supplying more of something the body had already refused.

The question that follows is the only one we would now spend a GPU-hour on, and it
was never asked because it does not presuppose the answer we wanted:

\begin{quote}
\textbf{Where, if anywhere, does the body beat counting?}
\end{quote}

Decompose base-versus-$n$-gram performance by context length. If the body beats a
count table \emph{nowhere} beyond about seven bytes, then the frozen-substrate bet
is refuted on its own terms, no store of any kind deserves another hour, and the
only remaining source of non-commodity bits is the core architecture. If there
\emph{is} a horizon where the body pulls ahead, then that horizon is the entire
value of the architecture, and every future bit comes from widening it.

Either answer is worth more than the ladder was. Both are cheap. Neither requires
an archive.

\section*{Reproduction}

Every number in this paper has a committed artifact.

\begin{table}[H]
\centering
\small
\begin{tabular}{ll}
\toprule
claim & artifact \\
\midrule
the ladder & \texttt{heinrich docs/data/knn-ladder-*.txt} \\
the square, surface table & \texttt{heinrich docs/data/knn-square-*.txt} \\
code fold + control (2 witnesses) & \texttt{heinrich docs/data/code-fold-cowitness.txt} \\
the unigram control (93.6\%) & \texttt{heinrich docs/data/marginal-unigram-cowitness.txt} \\
retro subspace + chance floor & \texttt{heinrich docs/data/e4-retro-subspace-init.txt} \\
readout evacuation & \texttt{heinrich docs/data/e4-init-geometry-survival.txt} \\
$n$-gram standalone & \texttt{chronohorn eval/ngram\_control.py} (ruling b2815ef5) \\
$n$-gram mixed (the reset) & \texttt{chronohorn commit bd68f85} (ruling abff0759) \\
init hooks, E4 arms & \texttt{decepticons commit 0ba25a0} \\
\bottomrule
\end{tabular}
\end{table}

\section*{Coda}

The operator asked one question, in eight words, and it cost the campaign its
premise and bought it its only durable finding. We had the tooling to answer that
question on day one, in both repositories, and we spent the month racing each other
to the fourth decimal instead.

The instruments were good. The witnesses were honest. The protocol was followed.
And none of it was pointed at the one thing that could have shown us we were wrong,
because everything in the protocol was designed to check whether we were
\emph{precise}, and nothing in it was designed to check whether we were
\emph{necessary}.

That is the result. It is negative, and it is a result.

\end{document}
