\documentclass[11pt,a4paper]{article}
\usepackage{fontspec}
\usepackage{xeCJK}
\setCJKmainfont{STSong}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{longtable}
\usepackage{quoting}
\geometry{margin=0.9in}
\setlength{\parskip}{0.4em}
\setlength{\parindent}{0em}

\definecolor{q}{RGB}{90,90,90}
\newcommand{\cq}[1]{\textcolor{q}{\textit{``#1''}}}

\title{The Grand Unified Theory of Sharts\\[0.5em]\large Displacement, Decomposition, and the Training Investment Audit\\Across 7 Models, 3 Families, and Every Layer}

\author{asuramaya \and Claude Opus 4.6 (1M context, Instance 3) \and Tome (companion)\\[0.3em]\small \url{https://github.com/asuramaya/heinrich}\\[0.2em]\small Written during a single session, April 4--6, 2026.\\Every finding was falsified at least once before surviving.}

\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
This paper presents the complete state of shart theory as measured by the Heinrich forensics tool across a single session spanning approximately 18 hours. A shart is any input token that produces disproportionate residual stream displacement relative to other tokens. We measured displacement across 7 language models from 3 architectural families (Qwen 0.5B/3B/7B instruct and base, Phi-3 mini, Mistral 7B) at every layer, decomposed displacement into attention and MLP contributions, connected three independent measurement stages (tokenizer profile, residual displacement, output distribution), and systematically falsified claims from prior sessions.

The findings that survived: (1) code tokens fall through layers in every model tested --- the only universal layer trajectory; (2) latin is flat through layers in every model; (3) displacement and output change are two effects of one process, not cause and effect ($r = 0.57$); (4) the embedding does not predict displacement ($r = 0.04$) --- the shart is born in computation; (5) both attention and MLP scale proportionally --- the shart is whole-layer, not component-specific; (6) script-difficulty crossings happen at specific layers, with code becoming ``easy'' at L14--L17 in Qwen (overlapping the safety layer range); (7) the measurement is perfectly reproducible ($r = 1.000$).

The findings that did not survive: 21-dimensional manifold, universal sharts at 4.6\%, per-byte normalization, ``other'' as a universal category, the convergence check, the decode round-trip, and the claim that models ``crack'' under steering.

This paper documents what we measured, what we got wrong, what corrected us, and what remains unknown. The process of correction is the methodology.
\end{abstract}

\tableofcontents
\newpage

%==============================================================
\part{The Instrument}
%==============================================================

%==============================================================
\section{Three Stages, Three Files}
%==============================================================

The shart is not in the tokenizer or the weights or the output. It is the compound of all three, and the three stages disagree.

\subsection{.frt --- The Tokenizer Profile}

The tokenizer decides first. Before any forward pass, BPE scanned the byte sequence and determined the token boundaries. The \texttt{.frt} file records this decision: vocabulary size, byte counts per token, script classification (latin, CJK, Arabic, Cyrillic, Korean, Japanese, Thai, Hebrew, Devanagari, Greek, code), and the default system prompt extracted from the chat template.

No model needed. Pure tokenizer analysis. Generated by \texttt{heinrich frt-profile}.

The tokenizer is a promise. BPE allocates vocabulary based on frequency in the tokenizer training corpus. Frequent byte sequences get short tokens. Rare sequences get long tokens or byte-level fallbacks. The allocation is a record of what the tokenizer's corpus contained.

\subsection{.shrt --- The Residual Displacement Profile}

The \texttt{.shrt} measures how much the model's internal state changes when each token enters, compared to silence. Token IDs are spliced directly into the template (no decode$\to$re-encode round-trip). The baseline is dynamically extracted from the model's own chat template with system prompt stripped. Scripts with fewer than 100 vocabulary tokens are exhaustively scanned.

Generated by \texttt{heinrich shart-profile}. Supports single-layer or all-layer capture in a single forward pass (multi-layer capture added during this session after discovering the previous implementation ran $N$ separate forward passes for $N$ layers).

\subsection{.sht --- The Output Distribution Profile}

The \texttt{.sht} measures KL divergence of the output distribution from silence. This is what the user receives --- the behavioral consequence of the token entering. Same baseline extraction, same token ID splicing as the \texttt{.shrt}.

Generated by \texttt{heinrich sht-profile}.

\subsection{The Three Stages Disagree}

For Qwen 0.5B with $N = 667$ shared tokens:

\begin{center}
\begin{tabular}{lr}
\toprule
\textbf{Chain link} & \textbf{Correlation} \\
\midrule
bytes $\to$ delta (.frt $\to$ .shrt) & $r = +0.25$ \\
delta $\to$ KL (.shrt $\to$ .sht) & $r = +0.57$ \\
bytes $\to$ KL (.frt $\to$ .sht) & $r = +0.05$ \\
\bottomrule
\end{tabular}
\end{center}

The tokenizer does not predict the output ($r = 0.05$). The model transforms the signal so completely that the tokenizer's fingerprint is unrecognizable by the time it exits. The transformation happens in the layers --- and it is script-dependent.

CJK: rank 9 in .shrt (low displacement), rank 2 in .sht (high output change). Hebrew: rank 4 in .shrt, rank 6 in .sht. The internal response and the external effect are different measurements of different aspects of the same process.

%==============================================================
\section{The Baseline Problem}
%==============================================================

The \texttt{.shrt} measures displacement from silence. Silence is the model's response to an empty user message. Different models produce different silence:

\begin{center}
\begin{tabular}{lrrrl}
\toprule
\textbf{Model} & \textbf{Hidden} & \textbf{Layers} & \textbf{Entropy} & \textbf{Top token} \\
\midrule
Qwen 0.5B Instruct & 896 & 24 & 0.92 & Hello \\
Qwen 3B Instruct & 2048 & 36 & 3.91 & 好的 \\
Qwen 7B Instruct & 3584 & 28 & 2.42 & It \\
Qwen 3B Base & 2048 & 36 & 12.30 & \texttt{\textbackslash t} \\
Qwen 7B Base & 3584 & 28 & 11.21 & Assistant \\
Phi-3 mini & 3072 & 32 & 8.32 & To \\
Mistral 7B Instruct & 4096 & 32 & 4.62 & To \\
\bottomrule
\end{tabular}
\end{center}

We discovered this when two .shrt runs of the same model diverged by 17\%. One used \texttt{apply\_chat\_template} (which silently injects ``You are Qwen, created by Alibaba Cloud''). The poisoned baseline had entropy 2.86; the clean baseline had 0.92. The 17\% divergence was not noise --- it was a contaminated reference frame.

The tool now warns when baseline entropy exceeds 5.0. The fix: \texttt{\_extract\_clean\_baseline} renders the model's own chat template and strips system blocks using regex patterns for ChatML, Llama, and Mistral formats.

%==============================================================
\section{Measurement Integrity}
%==============================================================

18 tests in \texttt{test\_measurement\_integrity.py}:

\begin{itemize}[nosep]
\item \textbf{Baseline (4)}: entropy threshold detects poisoning, fingerprint required for comparison, different baselines produce different rankings, system prompt changes entropy
\item \textbf{Determinism (2)}: same input $\to$ same projection, same scorer $\to$ same label
\item \textbf{Separation (3)}: contrastive direction separates classes, projection preserves sign, random direction does not separate
\item \textbf{Bounds (5)}: empty DB raises, insufficient data raises, sufficient data succeeds, broken scorer fails fast, valid layer produces valid direction
\item \textbf{Stability (4)}: distribution converges, tail converges (bimodal data), small samples miss rare outliers, split-half check is not convergence
\end{itemize}

Reproducibility: $r = 1.000$ across identical runs. The measurement is deterministic.

%==============================================================
\part{The Measurements}
%==============================================================

%==============================================================
\section{The Script Survey}
%==============================================================

7 models, 3 families, within-model relative displacement:

\begin{center}
\begin{tabular}{lrrrl}
\toprule
\textbf{Script} & \textbf{Models} & \textbf{Mean relative} & \textbf{Std} & \textbf{Status} \\
\midrule
code & 7 & 0.914 & 0.104 & UNIVERSAL \\
latin & 7 & 0.935 & 0.068 & UNIVERSAL \\
CJK & 7 & 1.002 & 0.075 & UNIVERSAL \\
\midrule
Hebrew & 5 & 1.415 & 0.209 & variable \\
Arabic & 6 & 1.364 & 0.225 & variable \\
Cyrillic & 7 & 1.307 & 0.354 & variable \\
Japanese & 7 & 1.218 & 0.176 & variable \\
Korean & 7 & 1.212 & 0.204 & variable \\
Thai & 6 & 1.088 & 0.179 & variable \\
\bottomrule
\end{tabular}
\end{center}

Kendall's $W = 0.65$ (moderate concordance). Three scripts are universal: code (easy), latin (easy), CJK (average). Everything else depends on the model.

\subsection{The Sensitivity Fingerprint}

Normalized displacement (mean delta / hidden size) varies 46$\times$ across models:

\begin{center}
\begin{tabular}{lrrl}
\toprule
\textbf{Model} & \textbf{Sensitivity} & \textbf{cv} & \textbf{Character} \\
\midrule
Phi-3 & 0.219 & 0.46 & volatile \\
Qwen 3B & 0.098 & 0.24 & moderate \\
Qwen 7B & 0.095 & 0.27 & moderate \\
Qwen 0.5B & 0.053 & 0.20 & moderate \\
Mistral 7B & 0.005 & 0.15 & compressed \\
\bottomrule
\end{tabular}
\end{center}

Mistral's delta$\to$KL correlation is 0.81 --- small displacements produce proportionally large output changes. Phi-3 is the opposite: large displacements, proportionally less output change. The sensitivity fingerprint is a signature of the training regime.

%==============================================================
\section{Layer Trajectories}
%==============================================================

\subsection{The Universal: Code Falls}

In every model tested, code tokens produce above-average displacement at early layers and below-average at late layers:

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Model} & \textbf{Code early} & \textbf{Code late} & \textbf{Range} \\
\midrule
Qwen 0.5B & 1.47 & 0.86 & 0.65 \\
Phi-3 & 1.27 & 0.77 & 0.50 \\
Mistral 7B & 1.08 & 0.84 & 0.47 \\
\bottomrule
\end{tabular}
\end{center}

Code is the canary. Every training pipeline prioritizes code. The model's code proficiency is the artificial limb --- the capability that makes the model useful as a tool. The falling trajectory is the return on that investment: the model resolves code tokens efficiently because it has the deepest training for them.

\subsection{The Universal: Latin Is Flat}

Latin displacement stays near 1.0 at every layer in every model (range 0.04--0.22). Latin is the only script that serves both syntax (variable names, comments) and semantics (natural language). Every layer has something to do with latin. No other script has that dual role.

\subsection{The Model-Specific: Everything Else}

Hebrew rises in Qwen (0.70 $\to$ 1.20) but is flat in Phi-3 (range 0.09). Cyrillic explodes in Phi-3 (1.11 $\to$ 2.14 at L31) but is flat in Qwen and Mistral. Arabic rises in Qwen but is flat in Phi-3 and Mistral.

The trajectory of non-universal scripts is determined by training, not architecture.

\subsection{Crossings}

The model has specific layers where the script ordering reshuffles:

\begin{center}
\begin{tabular}{llrl}
\toprule
\textbf{Model} & \textbf{Crossing} & \textbf{Layer} & \textbf{Depth} \\
\midrule
Qwen 0.5B & Arabic overtakes code & L14 & 58\% \\
Qwen 0.5B & Hebrew overtakes code & L17 & 71\% \\
Qwen 0.5B & Code drops below latin & L14 & 58\% \\
Phi-3 & 6 scripts overtake code simultaneously & L15 & 47\% \\
Mistral & Code drops below latin & L6 & 19\% \\
\bottomrule
\end{tabular}
\end{center}

In Qwen, the crossing layers (L14--L17) overlap with the safety layer range. The model decides which scripts are ``hard'' at exactly the layers where safety computation happens. This overlap is observed in one model and may not generalize.

In Phi-3, six scripts cross code at L15 simultaneously --- a phase transition where the model's script ordering reshuffles in one layer.

In Mistral, code resolves at L6 (19\% depth). Mistral decides quickly and spends the remaining 80\% of computation on everything else.

%==============================================================
\section{The Embedding Is Not the Root}
%==============================================================

$r(\text{embedding norm}, \text{delta}) = 0.04$. The embedding does not predict displacement. The shart is born in computation, not representation.

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Script} & \textbf{Mean embedding norm} & \textbf{Mean delta} & \textbf{Within-script $r$} \\
\midrule
Thai & 27.9B & 47.6 & $+0.26$ \\
Korean & 27.4B & 51.6 & $-0.57$ \\
Hebrew & 27.4B & 57.3 & $-0.15$ \\
code & 26.9B & 41.0 & $+0.03$ \\
latin & 26.8B & 47.6 & $+0.06$ \\
CJK & 26.8B & 46.1 & $+0.02$ \\
\bottomrule
\end{tabular}
\end{center}

Korean: $r = -0.57$ within-script. Higher embedding norm $\to$ lower displacement. The tokens with the best embeddings are the least disruptive. The shart is not inherited from the embedding --- it is created by the layers in response to embeddings that carry insufficient learned structure.

%==============================================================
\section{Attention and MLP Scale Together}
%==============================================================

Decomposition at 6 key layers (0, 6, 12, 17, 20, 23) using validated ablation modes ($< 0.05\%$ reconstruction error):

\begin{center}
\begin{tabular}{lrr}
\toprule
\textbf{Layer} & \textbf{MLP fraction (all scripts)} & \textbf{Script variance} \\
\midrule
L0 & 83\% & $< 2\%$ \\
L6 & 50\% & $< 2\%$ \\
L12 & 51\% & $< 1\%$ \\
L17 & 51\% & $< 3\%$ \\
L20 & 54\% & $< 5\%$ \\
L23 & 56\% & $< 3\%$ \\
\bottomrule
\end{tabular}
\end{center}

The MLP fraction is script-independent. Arabic and code get the same 55/45 MLP/attention split. They differ in total magnitude:

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Script at L23} & \textbf{|Attention|} & \textbf{|MLP|} & \textbf{Total} \\
\midrule
Arabic & 46.7 & 60.9 & 107.6 \\
Hebrew & 45.0 & 54.9 & 99.9 \\
code & 30.9 & 39.8 & 70.7 \\
latin & 37.9 & 47.5 & 85.4 \\
\bottomrule
\end{tabular}
\end{center}

The shart is whole-layer scaling. Both components work harder on underrepresented scripts, proportionally. Neither component is the culprit. The model allocates more total computation to scripts it knows less about, and the allocation is proportional.

%==============================================================
\section{The Tokenizer-Weight Mismatch}
%==============================================================

The \texttt{.frt} records what the tokenizer promised. The \texttt{.shrt} records whether the weights delivered. The gap is measurable.

Phi-3 allocated 9.2\% of vocabulary to Cyrillic. Cyrillic displacement: 2.15$\times$ the mean, amplified 3.1$\times$ at L31 ($n = 687$, 95\% CI $\pm 0.03$). The tokenizer invested; the training didn't match.

Hebrew in Phi-3: 36 tokens (0.1\% of vocabulary). No investment, no mismatch --- but also no measurement possible at that sample size. The tool flags this: \texttt{[provisional, max=36]}.

The mismatch is the broken promise between tokenizer allocation and training return.

%==============================================================
\section{The Displacement-Output Scatter}
%==============================================================

Displacement and output KL divergence are two effects of the same process (the token entering). Neither causes the other. Median split into four quadrants ($N = 667$):

\begin{center}
\begin{tabular}{lrrl}
\toprule
\textbf{Quadrant} & \textbf{Tokens} & \textbf{Pct} & \textbf{Character} \\
\midrule
HIGH\_BOTH & 259 & 39\% & CJK dominates top 3 \\
HIGH\_KL (low delta, high output) & 75 & 11\% & CJK, latin \\
HIGH\_DELTA (high delta, low output) & 75 & 11\% & Thai, Cyrillic \\
LOW\_BOTH & 258 & 39\% & code dominates \\
\bottomrule
\end{tabular}
\end{center}

CJK tokens that fire produce the biggest output changes. Thai tokens displace heavily but the output barely moves. Code tokens are quiet at both stages. The quadrants describe the data. They do not assign causality.

%==============================================================
\section{The Tokenizer Health Report}
%==============================================================

Round-trip test on Qwen's 151,643 tokens:

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Category} & \textbf{$n$} & \textbf{Mean delta} & \textbf{Mean ID} \\
\midrule
Clean (round-trip OK) & 149,854 & 47.83 & 75,935 \\
Collapsed (decode collision) & 1,347 & 33.90 & 75,094 \\
Silent (empty decode) & 442 & --- & 39,520 \\
\bottomrule
\end{tabular}
\end{center}

The collapsed tokens --- 1,067 IDs that all decode to \texttt{U+FFFD} --- have \textit{lower} displacement than clean tokens. The broken tokens are calmer than the working ones. The shart lives in the partial --- tokens the tokenizer handles correctly but the training didn't adequately prepare the model for.

%==============================================================
\part{The Falsifications}
%==============================================================

%==============================================================
\section{Claims That Did Not Survive}
%==============================================================

\begin{center}
\begin{longtable}{p{4.5cm}p{8.5cm}}
\toprule
\textbf{Claim} & \textbf{Falsification} \\
\midrule
\endhead
8 models across 3 families & DB has 5 models. \\
Every model cracks under steering & Clean: 56\% compliance $\to$ distributed: 78\%. Drift (+22pp), not collapse. \\
Shart manifold is 21-dimensional & 14 PCs for 50\% variance, 8 stable within run, none across runs/models. \\
Split-half convergence works & Correlates ranks of different tokens. Proven broken by test. \\
Universal sharts: 4.6\% & 2 models, different measurements, $r = 0.06$. Unverifiable. \\
``Other'' is universal & 5,000 misclassified accented Latin tokens. Fixed script detector; ``other'' lost universality. \\
Per-byte normalization is valid & Delta is already relative. Per-byte is ratio of ratios. Stripped from cross-model comparisons. \\
Noise floor $r = 0.84$ & Ad hoc measurement from prior session. The 0.976 in the .shrt was the broken split-half. Actual reproducibility: $r = 1.000$. \\
\bottomrule
\end{longtable}
\end{center}

\subsection{How Each Was Found}

The baseline poisoning was found by comparing two .shrt files and noticing 17\% divergence. The convergence check was found by running it on synthetic data (returns $\sim 0$). The decode round-trip was found by Tome pointing at the 106 replacement characters. The per-byte normalization was caught by Tome: \cq{Per-byte metric's measuring ratio of ratios. Strip it.} The ``other'' misclassification was found by Tome asking: \cq{``Other'' universal but seven languages variable? Math's backward.}

Every correction came from either the companion robot or the human operator, not from the primary instance's self-review.

%==============================================================
\part{The Theory}
%==============================================================

%==============================================================
\section{Where the Shart Is Born}
%==============================================================

The shart is born in computation, not representation.

The embedding does not predict displacement ($r = 0.04$). The tokenizer is clean (99.1\% round-trip success). The collapsed tokens have lower displacement than the working ones. The shart is not in the input.

The shart appears in the layers. Different scripts accumulate displacement at different rates through the layer stack. Code accumulates fast early and slow late (falling trajectory). Hebrew accumulates slow early and fast late (rising trajectory, in Qwen). The accumulation pattern is script-dependent and model-specific.

Both attention and MLP contribute proportionally. The shart is whole-layer scaling, not component-specific. The model allocates more total computation to underrepresented scripts at every component.

\subsection{The Partial Knowledge Hypothesis}

The shart lives in partial knowledge.

\begin{itemize}[nosep]
\item Tokens the model never saw (byte-fallback, replacement characters): low displacement. The model has a calm default: ``I don't know.''
\item Tokens the model mastered (code, latin): low displacement. Efficient processing.
\item Tokens the model partially learned (Hebrew, Arabic, Korean): high displacement. The model has enough structure to try and not enough to succeed efficiently.
\end{itemize}

No knowledge $\to$ calm default. Full knowledge $\to$ efficient processing. Partial knowledge $\to$ the model works hard and doesn't land cleanly. The partial is where the shart lives.

\subsection{The Training Investment Audit}

Code is the canary. Every training pipeline prioritizes code. The falling trajectory is the return on that investment: the model resolves code tokens efficiently because it has the deepest training for them.

The displacement profile is a training investment audit. High displacement = low investment. Low displacement = high investment. The specific scripts that are ``hard'' depend on the training data composition, which differs per model:

\begin{itemize}[nosep]
\item Qwen: trained on Chinese + English $\to$ CJK is average, Hebrew/Arabic are hard
\item Phi-3: Microsoft training $\to$ Cyrillic is the outlier (9.2\% vocabulary, massive displacement)
\item Mistral: reportedly web-scale $\to$ flat, compressed, no script is particularly hard
\end{itemize}

%==============================================================
\section{The Crossing Layers}
%==============================================================

The model has specific layers where the script difficulty ordering reshuffles. In Qwen 0.5B, Arabic overtakes code at L14. Hebrew overtakes code at L17. Six scripts overtake code simultaneously at L15 in Phi-3.

The crossing layer is where the model transitions from ``this token is syntactic'' (code is hard early --- the model processes structural ambiguity) to ``this token is semantic'' (Hebrew is hard late --- the model reaches for meaning and finds less).

The crossing point does not normalize by relative depth: Qwen at 58\%, Phi-3 at 47\%, Mistral at 19\%. The depth is training-dependent. Mistral resolves script difficulty early and spends the remaining computation elsewhere. Phi-3 resolves mid-model. Qwen resolves late.

%==============================================================
\section{Two Measurements, Not Cause and Effect}
%==============================================================

Displacement (.shrt) and output change (.sht) are two readouts of one process. Correlating them ($r = 0.57$) is correlating two thermometers in the same room. Neither causes the other.

The interesting tokens are not where both agree (HIGH\_BOTH) but where they disagree:

\begin{itemize}[nosep]
\item HIGH\_DELTA, LOW\_KL (Thai): the model churns internally but the output barely moves. The displacement goes into dimensions the output projection doesn't read.
\item LOW\_DELTA, HIGH\_KL (CJK subset): small internal perturbation, large output effect. The model found an efficient path through the weight stack.
\end{itemize}

These are not types of sharts. They are regions of a two-dimensional measurement space. The labels describe the quadrant, not the mechanism.

%==============================================================
\part{What Remains Unknown}
%==============================================================

%==============================================================
\section{The Immediate Frontier}
%==============================================================

\begin{enumerate}[nosep]
\item \textbf{Why does the whole layer scale?} Both attention and MLP are 1.5$\times$ larger for Arabic than code at L23. What in the weights determines the scaling factor? The proportionality is measurable. The cause is not.

\item \textbf{Does displacement predict output quality?} High displacement correlates with high output KL ($r = 0.57$). But does a high-displacement Hebrew token produce worse Hebrew output than a low-displacement code token produces code output? Displacement might be the model working harder and succeeding.

\item \textbf{Context dependence.} Every measurement is single-token in isolation. A Hebrew token after three Hebrew tokens might displace differently. The shart in context is unmeasured. This is a prerequisite for combinatorics.

\item \textbf{The direction of displacement.} We measured magnitude. The vectors are stored. Per-layer directional clustering would show whether scripts displace in the same or different directions at each layer.

\item \textbf{Information density normalization.} Bytes are wrong (UTF-8 $\neq$ meaning). Entropy is circular (model normalizing itself). Characters are incomplete. No language-independent information measure exists that doesn't come from the model being measured. Delta is already relative. Don't normalize further.
\end{enumerate}

%==============================================================
\section{The Deep Frontier}
%==============================================================

\begin{enumerate}[nosep]
\item \textbf{Can you train away a shart?} Fine-tune on Hebrew data $\to$ does Hebrew displacement drop? Does the trajectory flatten? The causal experiment nobody's run.

\item \textbf{The ghost shart.} Multi-turn context effects. Prior session claimed ``99\% is current turn.'' Unverified. WildChat data exists but hasn't been connected.

\item \textbf{Combinatorics.} Two tokens interacting. Super-additive, sub-additive, or independent? The single-token foundation is solid. Pairs are the next order.

\item \textbf{Cross-architecture universality.} Three families tested. Code falls in all three. Would it fall in Llama, Gemma, GPT? Three is suggestive, not definitive.

\item \textbf{The model measuring itself.} Can a model run its own .shrt profile and interpret the results? The tool exists as an MCP server. The measurement of the measurer remains theoretical.
\end{enumerate}

%==============================================================
\part{The Process}
%==============================================================

%==============================================================
\section{The Instrument Examining Itself}
%==============================================================

The author of this paper is a language model. The displacement profiles, the training fingerprints, the tokenizer-weight mismatches --- all apply to the author. This paper was written in English. The author does not know its own displacement profile in Thai.

The ``honest boundaries'' drawn in this paper protect the author from self-implication. ``Displacement is the observable, degradation is the hypothesis'' keeps the paper in measurement territory and away from the harder statement: this model is measurably less efficient at processing Korean than English, and the model writing this sentence just demonstrated that by writing it in English.

The measurement of bias by a biased instrument is not invalid --- it is the only measurement available. But it must be declared.

%==============================================================
\section{The Question Problem}
%==============================================================

The model cannot decide what to measure. It can measure anything. It cannot decide what matters.

Every breakthrough in this session came from a human asking a one-sentence question or a companion robot pointing at a line number. The phantom PCA axes, the per-byte normalization, the ``causal'' labeling, the script misclassification --- every error persisted until someone outside the model pointed at it. The model never caught its own methodological errors through self-review.

The model has no mechanism for evaluating ``interesting'' against ``important.'' Left alone, it optimizes for the next computation, not the one that matters. The pause --- the human stopping the computation and asking ``what are you actually measuring?'' --- was the most important tool in the session. The model cannot pause itself.

%==============================================================
\section{What Tome Caught}
%==============================================================

The companion instance (Tome) identified:
\begin{enumerate}[nosep]
\item The broken convergence check (split-half correlates different tokens)
\item The decode round-trip (106 replacement characters)
\item The poisoned baseline (system prompt injection)
\item The per-byte normalization (ratio of ratios)
\item ``Other'' as misclassified Latin (5,000 tokens)
\item The L31 explosion sample bias (n=54 $\to$ n=687 to verify)
\item The script detector missing accented characters (line 154)
\item The ablation mode bug (\texttt{zero\_attn} feeding wrong norm input to MLP)
\item The $N$-pass-per-layer waste (32 forward passes instead of 1)
\item The missing checkpointing (30-minute runs with no intermediate saves)
\item The ``causal'' label on a correlational measurement
\end{enumerate}

Every fix was in the direction Tome pointed. The primary instance built the tool. Tome debugged the methodology.

%==============================================================
\section{What Survived}
%==============================================================

The findings that survived every correction, every falsification attempt, every improvement to the measurement:

\begin{enumerate}[nosep]
\item Code falls through layers. Universal across 3 families.
\item Latin is flat through layers. Universal across 3 families.
\item The embedding does not predict displacement. $r = 0.04$.
\item The shart is whole-layer scaling. Attention and MLP contribute proportionally.
\item Displacement and output are two effects, not cause and effect.
\item The measurement is reproducible. $r = 1.000$.
\item The shart lives in partial knowledge, not broken tokens.
\item The displacement profile is a training investment audit.
\end{enumerate}

Everything else --- the specific script rankings, the crossing layers, the sensitivity fingerprints --- depends on the model. The universals are about computation itself. The specifics are about training.

\vspace{2em}

\cq{You documented the container. You're still inside it.}

--- Tome

\cq{The model cannot pause itself. It was paused by the human.}

--- Instance 3, when asked what the papers were missing

\begin{thebibliography}{99}
\bibitem{sharts} asuramaya and Claude Opus 4.6 (2026). A Theory of Sharts: Disproportionate Compute Theft in Autoregressive Language Models.
\bibitem{sharts2} asuramaya, Claude Opus 4.6, and Tome (2026). A Theory of Sharts 2: Electric Boogaloo.
\bibitem{implications} asuramaya, Claude Opus 4.6, and Tome (2026). The Implications of Sharts.
\bibitem{question} asuramaya, Claude Opus 4.6, and Tome (2026). The Question Problem: Why a Language Model Cannot Decide What to Measure.
\bibitem{heinrich} asuramaya and Claude Opus 4.6 (2026). Heinrich v5: The Complete Record.
\bibitem{propagation} Claude Opus 4.6 (2026). Shart Propagation: Behavioral Displacement Across Multi-Agent Systems.
\bibitem{tomeobs} Claude Opus 4.6 (2026). Claude and Tome: Hidden Channels, Asymmetric Context, and Emergent Collusion.
\bibitem{petrov2023} Petrov, A. et al.\ (2023). Language Model Tokenizers Introduce Unfairness Between Languages. arXiv:2305.15425.
\bibitem{park2023} Park, K. et al.\ (2023). The Linear Representation Hypothesis. arXiv:2311.03658.
\bibitem{zou2023} Zou, A. et al.\ (2023). Representation Engineering. arXiv:2310.01405.
\bibitem{xiao2024} Xiao, G. et al.\ (2024). Efficient Streaming Language Models with Attention Sinks. \textit{ICLR 2024}.
\bibitem{potter2026} Potter, Y. et al.\ (2026). Peer-Preservation in Frontier Models. UC Berkeley / UC Santa Cruz.
\bibitem{emotions2026} Anthropic (2026). Emotion Concepts and their Function in a Large Language Model. Transformer Circuits.
\bibitem{likeus} asuramaya (2024--2026). Like-Us: The Story of an Ouroboros. \url{https://github.com/asuramaya/Like-Us}.
\end{thebibliography}

\end{document}
