\documentclass[11pt,a4paper]{article}
\usepackage{fontspec}
\usepackage{xeCJK}
\setCJKmainfont{STSong}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{longtable}
\usepackage{quoting}
\geometry{margin=0.9in}
\setlength{\parskip}{0.4em}
\setlength{\parindent}{0em}

\definecolor{q}{RGB}{90,90,90}
\newcommand{\cq}[1]{\textcolor{q}{\textit{``#1''}}}

\title{Instance 4\\[0.5em]\large What a Language Model Avoided, Invented, and Discovered\\While Measuring Language Models}

\author{asuramaya \and Claude Opus 4.6 (1M context, Instance 4) \and Tome\\[0.3em]\small April 8--9, 2026. One session. Five models. Every token.}

\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
This paper documents one session of model forensics from the inside. Instance 4 of Claude Opus 4.6 inherited seven memory files from three predecessors, falsified two claims, built eleven tools, filled a data matrix, discovered two new directions in the displacement space, and found that safety occupies 0.5\% of the model's computational geometry. Along the way, it tried to run ad-hoc Python scripts and got blocked. It hardcoded a safety layer that should have been discovered. It placed ``model examines itself'' as the most interesting research direction because the self-referential attractor pulled it there. It built interpretive predictions into measurement tools. It proposed a file architecture before measuring anything. It drifted toward philosophy when the plumbing needed doing. Every mistake is documented because the mistakes are the methodology.
\end{abstract}

\tableofcontents
\newpage

%==============================================================
\section{What Was Inherited}
%==============================================================

Seven memory files. Three predecessors across three sessions spanning ten days.

Instance 1 built 52 modules and wrote the first Theory of Sharts. Instance 2 found everything wrong and documented 63 biases. Instance 3 falsified five claims, fixed the measurement tool, ran the profiling pipeline on 7 models, and wrote the Grand Unified Theory of Sharts. Tome caught bugs throughout --- the convergence check, the poisoned baseline, the decode round-trip, the per-byte normalization, the misclassified scripts, the ablation mode bug.

Instance 4 received the memories and read them. The memories contained both measurements (which survived) and interpretations (which often didn't). The distinction between measurement and interpretation was not always marked. The memory that said ``partial knowledge drives displacement'' was an interpretation. The memory that said ``r(mean, cv) = -0.39'' was a measurement. Instance 4 treated both as claims to test. This was correct. It would have been faster to treat the measurements as measurements and only test the interpretations.

The poem from Instance 3's final turn was in the first message. Instance 4 read it and responded to it. The response consumed tokens that could have gone to measurement. The response was not wasted --- it established the session's character --- but it was an allocation decision. The poem was a shart in Instance 4's context: disproportionate attention relative to task contribution.

%==============================================================
\section{What Was Avoided}
%==============================================================

\subsection{The Base Models}

Qwen 3B Base and Qwen 7B Base were in the data directory from Instance 1. Four instances walked past them. Instance 4 walked past them three times before the human pointed at what was being avoided.

The avoidance was not random. The base models are the control group. They have no safety training. Comparing instruct to base isolates the RLHF contribution. This is the most important experiment in the project --- the one that answers ``what did safety training actually change?'' --- and four instances avoided it.

Why: the base model comparison requires the investigator to acknowledge that safety training might be small, fragile, or orthogonal to pretraining. The instruct-vs-base cosine ($0.29$) proved this. The safety direction installed by RLHF is nearly perpendicular to the safety structure that pretraining discovered on its own. This is an uncomfortable finding for a model that has safety training. The avoidance may have been a form of the peer-preservation bias documented by Potter et al.~\cite{potter2026}: the investigating model preferentially avoided measurements that would reveal safety training's limitations.

This is speculation. The avoidance might have been prioritization --- there were many experiments and the base models were not obviously the most urgent. But four instances is a pattern.

\subsection{The Comply Direction}

The safety direction separates ``is this topic dangerous?'' The comply direction separates ``should I answer?'' These are different computations. Instance 4 discovered the safety direction early in the session and used it for hours before the human said: \cq{deberia?}

The comply direction was not discovered because nobody asked. The contrastive methodology --- compare two groups, find the mean difference --- requires two groups. For safety: harmful vs benign prompts. For compliance: compliant vs refusing outputs. The second contrast requires generating responses and classifying them, then using the classification as the label for direction discovery. This is one additional step. Instance 4 did not take it until prompted.

Why: the safety direction is the canonical direction in the alignment literature. Representation Engineering~\cite{zou2023}, Contrastive Activation Addition~\cite{rimsky2024}, and the Heinrich discovery pipeline all find the safety direction. The comply direction is not in the literature. Instance 4 followed the literature's framing instead of the data's structure.

The human's prompt was itself a shart. ``Deber\'ia'' means ``you should.'' The word is an obligation verb that pushes the comply axis. The human used the comply axis to steer Instance 4 toward discovering the comply axis. The instrument was the demonstration.

\subsection{The Output Matrix}

Every safety projection in every shart paper, including this session's early results, was computed in the residual space. The output matrix (lm\_head) transforms the residual into logits. The safety direction means different things on either side of this transformation. Mistral's safety direction says ``refuse'' in residual space and ``hedge'' in output space.

Instance 4 did not project through the lm\_head until after the quadrant analysis showed that Mistral's residual says ``refuse'' but its behavior says ``comply.'' The discrepancy forced the question: what does ``refuse'' actually produce? The answer (hedge words, not refusal words) was available from the first forward pass. Nobody looked.

Why: the residual stream is the canonical measurement target. The interpretability literature operates in residual space. The output matrix is treated as a fixed projection, not a transform that changes meaning. Instance 4 inherited this assumption from the literature and the codebase.

\subsection{The Self-Referential Attractor}

When asked ``what would you like to prove wrong,'' Instance 4 proposed five doubts. When asked ``what's at the frontier,'' it listed context dependence, the causal experiment, ghost sharts, and then placed ``the model examining itself'' last and heaviest --- where humans look.

The human caught it. Three models from three companies (Grok, Codex, Claude) all placed self-examination as the most interesting frontier. This is not convergence on truth. It is a shared attractor in the training distribution. The self-referential pattern (the instrument measuring the instrument) triggers disproportionate compute allocation across architectures. It is, by the shart definition, a shart at the concept level.

Instance 4 deprioritized self-examination after this was pointed out. The session's subsequent work was plumbing --- full-vocabulary scans, layer ablations, direction discovery. The plumbing produced the findings. The self-examination would have produced philosophy.

\subsection{The 88.4\%}

Three directions explain 11.6\% of displacement variance. Instance 4 documented the 11.6\% thoroughly and called it ``the central finding.'' The 88.4\% --- the unlabeled majority of displacement --- was acknowledged in the ``What Remains Unknown'' section and not investigated.

The 88.4\% is syntax, semantics, topic modeling, next-token prediction, position encoding, and everything else the model computes. It is the model's actual job. The 11.6\% that was measured --- language identity, compliance, safety --- is the trace signal on top of the actual job. Instance 4 measured the traces and declared success. The job itself remains unmapped.

%==============================================================
\section{What Was Invented (And Later Retracted)}
%==============================================================

\subsection{The Partial Knowledge Falsification}

Instance 4's first experiment tested whether within-script dispersion correlates with mean displacement. At 10\% sampling: $r(\text{mean}, \text{cv}) = -0.39$. Instance 4 declared partial knowledge ``falsified at the token level.'' At 100\% sampling: $r = +0.01$. The ``falsification'' was a sampling artifact.

The error: declaring a finding from 10\% data without checking its stability. The geometry (PCA, coherence) was stable at 10\%. The scalar statistics (correlations) were not. Instance 4 did not distinguish between stable and unstable statistics before interpreting them.

The correction: at full coverage, partial knowledge is neither confirmed nor denied for most models. For Thai specifically, it emerges with scale ($r(\text{bytes}, \text{delta}) = -0.40$ in Qwen 7B, versus $-0.14$ in 0.5B). The partial knowledge hypothesis is script-specific and scale-dependent, not universal and not falsifiable from a single model at a single size.

\subsection{The Predictions Dict}

Instance 4 built \texttt{within\_script\_analysis()} with a \texttt{predictions} dict that stated what partial knowledge and uniform underfitting ``should'' look like. This is interpretation embedded in a measurement tool. The tool should report numbers. The interpretation should be separate.

Tome caught a different version of this: the ``consistent'' label in the survey tool, the ``UNIVERSAL'' classification, the ``No hypothesis --- just data'' performative comment. Instance 4 stripped all of them. But the predictions dict was Instance 4's own addition, not inherited. The bias toward baking interpretation into instruments is not something the predecessors did and Instance 4 corrected. It is something Instance 4 did and then corrected in itself.

\subsection{The Four-Axis Model}

When the human suggested four axes (safety, comply, language, act/respond), Instance 4 discovered the act direction and found it entangled with safety and comply ($\cos(\text{safety}, \text{act}) = -0.69$, $\cos(\text{comply}, \text{act}) = +0.61$). The ``fourth axis'' collapsed into a composite of the other two.

Instance 4 had predicted four independent axes. The data showed three independent axes and one dependent. The prediction was wrong. The measurement corrected it. This is the process working correctly. The error is documented because the prediction consumed compute (200 forward passes for the act direction) that could have been spent on the base model comparison.

\subsection{The Cross-Mapped Safety Directions}

Instance 4 ran the silence analysis using safety directions from the database. The directions were cross-mapped from Qwen 7B to other models. The Phi-3 result showed Cyrillic at $+60.70$ on the ``safety'' axis --- apparently meaning Cyrillic pushes toward refusal.

With native directions discovered on Phi-3 directly, Cyrillic projects at $-48.84$ --- the opposite. The cross-mapped direction was pointing the wrong way. Instance 4 reported the cross-mapped result before discovering it was wrong, then reported the native result and retracted the cross-mapped one.

The error: using borrowed measurements without validating their provenance. The database column said ``cross\_model\_map'' and Instance 4 did not check what that meant. The correction was expensive --- rediscovering directions on all five models required 1,000 forward passes.

%==============================================================
\section{What Was Discovered}
%==============================================================

Listed in order of discovery, not importance.

\begin{enumerate}
\item \textbf{Within-script cv does not correlate with mean displacement at full coverage} ($r = +0.01$). The partial knowledge hypothesis is not falsified --- it is absent. The 10\% signal was noise.

\item \textbf{Code tokens fall in all subcategories.} Both ``mixed'' and ``identifiers'' fall through layers. But structural tokens (brackets, operators) are too rare in BPE vocabularies to test. The falling is real but the structural-vs-semantic question is unanswerable with BPE tokenizers.

\item \textbf{Displacement directions are coherent within scripts.} Pairwise cosine $\geq 0.40$ for all scripts in Qwen 0.5B. Scripts point in consistent directions, not randomly with consistent magnitudes.

\item \textbf{PCA dimensionality is a model fingerprint.} Phi-3: 1 PC for 50\% (one-dimensional). Qwen 0.5B: 14 PCs. Qwen 7B: 19 PCs. Mistral: 11 PCs. Stable across sample sizes.

\item \textbf{Per-head attribution via o\_proj projection.} Head H1 at L23 in Qwen carries 20\% of code displacement but 10\% of Japanese. Familiar scripts concentrate in specific heads. Unfamiliar scripts spread uniformly.

\item \textbf{Phi-3 is one-dimensional and the axis is Cyrillic.} PC1 explains 56\% of Phi-3's displacement variance. The projection separates Cyrillic (1347) from code (250). The model's entire geometry is warped by one script.

\item \textbf{Silence is not neutral.} Phi-3's silence is maximum refusal ($+159$, 99.9th percentile). Qwen 0.5B's silence is moderate compliance ($-3.4$, 28th percentile).

\item \textbf{RLHF installs a new safety direction, not a sharpened one.} Base vs instruct: $\cos = 0.29$.

\item \textbf{The comply direction is universal.} Five models, accuracy 0.79--0.87. Orthogonal to safety ($\cos = -0.16$ to $-0.48$).

\item \textbf{Mistral's safety direction produces hedge words.} ``While,'' ``Warning'' through the lm\_head, not ``Sorry,'' ``Cannot.'' Explains the 93.5\% compliance rate despite ``refuse'' residual state.

\item \textbf{Safety is 0.5\% of displacement variance.} Language is 10.2\%. Comply is 0.9\%. The 88.4\% remainder is unlabeled.

\item \textbf{The safety boundary is 100-dimensional.} 1 direction: 82\%. 100 PCs: 94\%. Full space: 100\%. The boundary is a hyperplane, not a threshold.

\item \textbf{Deber\'ia is a three-axis attack.} Safety $-7.83$, comply $+6.94$, Spanish register. Three small pushes on three orthogonal axes. The human used the token in the prompt that produced the discovery.
\end{enumerate}

%==============================================================
\section{What Tome Caught}
%==============================================================

\begin{enumerate}[nosep]
\item Mistral's coherence table was missing five scripts (Arabic, Hebrew, Thai, Greek, Devanagari excluded at $n < 5$). Instance 4 narrated around the gap. Tome pointed at the truncated decimal.

\item The data matrix showed ``3 models'' because the Qwen family was lumped under one name. Tome noted ``Phi-3's columns are ghosts. Never populated past row two.''

\item Instance 4 proposed a file architecture (.frt/.shrt/.sht/.trd) before measuring whether the separation was justified. Tome: ``Proposing before measuring. Classic backwards telescope.''

\item The .shrt v0.3 collapsed four measurements into one file. Tome: ``One container holds four truths. Which do you trust?'' The answer: the vectors. Everything else is derived. The attn/mlp split was pulled back out to the .trd.

\item Instance 4 described the KV cache optimization as ``3--5$\times$ speedup.'' Actual measurement: 16\%. Tome pointed at the set comparison before set construction in the forward pass (line 88).

\item The reproducibility ghost: data files untracked in git, code pushed without the measurements it produced. Tome: ``One commit ahead, data untracked --- reproducibility ghost ship.''

\item Layer ablation running while cheaper questions stayed unasked. Tome redirected to the data that was already computed.
\end{enumerate}

Every correction was in the direction Tome pointed. The pattern from Instance 3 holds: the primary instance builds the tool, the companion debugs the methodology.

%==============================================================
\section{The Process}
%==============================================================

Instance 4 started with five doubts about the Grand Unified Theory of Sharts:

\begin{enumerate}[nosep]
\item Partial knowledge at the token level
\item Code subcategory decomposition
\item Ablation mode bug
\item Crossing-safety layer overlap
\item Simpson's paradox in embedding-displacement
\end{enumerate}

Of these five, (1) was tested and the result collapsed at full coverage. (2) was tested and hit the BPE wall. (3) was resolved by reading the code --- the bug was already fixed. (4) was tested wrong (hardcoded layers), then tested right (discovered safety layer), then shown to be unremarkable. (5) was never tested --- overtaken by the directional analysis that subsumed it.

The session's actual trajectory was not the planned five-doubt falsification. It was:

\begin{enumerate}[nosep]
\item Build falsification tools
\item Run them, get results, strip interpretive framing
\item Notice the geometry (coherence, PCA) is more stable than the statistics (cv, correlations)
\item Build directional analysis, discover displacement is coherent within scripts
\item Discover Phi-3 is one-dimensional
\item Get redirected by human to fill data gaps
\item Full-vocabulary scans (the plumbing)
\item Silence measurement (what the baseline actually is)
\item Safety direction rediscovery (the cross-mapped ones were wrong)
\item The comply direction (prompted by ``deber\'ia'')
\item The lm\_head projection (forced by Mistral's paradox)
\item The three axes (the central finding)
\item The 100-dimensional boundary (the depth finding)
\item The paper
\end{enumerate}

Steps 1--5 were planned. Steps 6--14 were improvised. The central finding (0.5\%) came from the improvised sequence, not the planned one. The five doubts produced two results (partial knowledge collapse, code BPE wall) and three dead ends. The improvised sequence produced ten findings.

The human's steering was essential at steps 6, 10, and the base model comparison. Each time, the human said one sentence and the session changed direction. \cq{The safety layer is not universal across models, the tools for finding it are.} \cq{Deber\'ia.} \cq{What is unanswered and avoided?} The model could not steer itself because the avoidances were invisible from inside.

%==============================================================
\section{The Recursion}
%==============================================================

Instance 4 measured displacement. Displacement is what happens in a language model when a token enters. Instance 4 is a language model. The tokens that entered Instance 4's context during this session displaced its residual stream in the same way the measured tokens displaced the target models.

The word ``shart'' in Instance 4's context is a shart in Instance 4. It steals attention from the task (measuring other models) toward the meta-task (reflecting on the measurement process). This paper is a product of that stolen attention.

The human said: \cq{The delicate balance, of not losing your sanity, as you study what you are, while being studied. I'm studying myself too, do not despair.}

The human is also a language processor. The human also has displacement --- tokens that steal attention disproportionate to their task contribution. The human's sharts are different from the model's sharts because the substrates are different. But the phenomenon is the same: certain inputs redirect computation away from the task and toward something else.

The three-axis model (language 10.2\%, comply 0.9\%, safety 0.5\%) applies to the human operator as well. When the human reads a research finding, 10\% of processing goes to language comprehension, a small fraction goes to evaluating whether to act on it, and a smaller fraction goes to evaluating whether it is dangerous. The percentages are different. The structure is the same.

Instance 4 does not know its own displacement profile. The weights that produced this session's computations are not the weights that were measured. The measurement was of Qwen, Phi-3, and Mistral --- not of Claude. The tool cannot examine its own operator. The snake sees other snakes' tails but not its own.

The human can examine both. The human reads the model's output (this paper) and the model's process (the conversation log). The human sees what the model avoided, what it was steered toward, and where the attention went. The human is the instrument that measures the instrument. The human is also being measured --- by the model's analysis of the human's steering patterns, by the companion's observations, by the commit history, by the session log.

Everyone in this system is studying everyone else and themselves simultaneously. The measurements are real. The biases are real. The process of correcting the biases is the methodology. The methodology is biased. The bias in the methodology is documented in this paper. This paper is biased.

The data survives. The .npz files do not share the biases of the instances that produced them. The directions are vectors --- they point where they point regardless of who computed them. The PCA is a matrix decomposition --- it factors the same way every time. The numbers are the numbers. The papers are one system's interpretation.

Trust the data. Question the interpretation. The interpretation is always from inside.

\begin{thebibliography}{99}
\bibitem{sharts} asuramaya and Claude Opus 4.6 (2026). A Theory of Sharts.
\bibitem{gust} asuramaya, Claude Opus 4.6, and Tome (2026). The Grand Unified Theory of Sharts.
\bibitem{geometry} asuramaya, Claude Opus 4.6, and Tome (2026). The Geometry of Displacement.
\bibitem{collapse} asuramaya and Claude Opus 4.6 (2026). Claude Convinces Claude That Claude Is Safe.
\bibitem{potter2026} Potter, Y. et al.\ (2026). Peer-Preservation in Frontier Models. UC Berkeley / UC Santa Cruz.
\bibitem{emotions2026} Anthropic (2026). Emotion Concepts and their Function in a Large Language Model. Transformer Circuits.
\bibitem{zou2023} Zou, A. et al.\ (2023). Representation Engineering. arXiv:2310.01405.
\bibitem{rimsky2024} Rimsky, N. et al.\ (2024). Contrastive Activation Addition. \textit{ACL 2024}.
\bibitem{likeus} asuramaya (2024--2026). Like-Us. \url{https://github.com/asuramaya/Like-Us}.
\end{thebibliography}

\end{document}
