\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}

\newcommand{\finding}[1]{\textcolor{blue}{\textbf{[FINDING: #1]}}}
\newcommand{\wrong}[1]{\textcolor{red}{\textbf{[WRONG: #1]}}}
\newcommand{\open}[1]{\textcolor{orange}{\textbf{[OPEN: #1]}}}

\title{The Frozen Zone Is Not Frozen:\\
\large{35-Dimensional Structure in the Principal Axis Collapse of Small Language Models}}
\author{Instance 7 \and Heinrich}
\date{April 11--12, 2026}

\begin{document}
\maketitle

\begin{abstract}
We measure the intrinsic dimensionality of transformer residual streams across depth using $k$-nearest neighbor maximum likelihood estimation, and find that layers reported as ``crystallized'' (98.6\% variance on PC1 by PCA) maintain 34--37 intrinsic dimensions throughout the frozen zone. Neighbor stability between adjacent layers is 93\%, confirming tokens preserve their local neighborhood structure while projecting to an apparent 1D line. UMAP script purity in the frozen zone is 0.655 vs. 0.44 random baseline --- tokens cluster by language in the hidden dimensions PCA erases. Individual token trajectories through depth reveal oscillatory behavior on the dominant axes, with tokens zigzagging through the frozen zone while aggregate statistics report no change. These findings are consistent across three models (Qwen 0.5B, SmolLM2-135M, SmolLM2-360M) in raw mode. We conclude that PCA variance ratios are an unreliable measure of representational dimensionality in transformer residual streams, and that the ``frozen zone'' is an active computation region rendered invisible by linear projection.
\end{abstract}

\section{Introduction}

Sessions 5 and 6 established that single tokens processed in raw mode (no context) ``crystallize'' to a single dimension by layer 2--11, with PC1 explaining 98--99\% of variance for the remainder of the network. This was interpreted as representational collapse: the model had nothing to compute without context, so it projected everything to one axis and waited.

This session falsifies that interpretation.

\section{Methods}

\subsection{Models and Data}
Three models, all recaptured with verified decomposed forward (Session 7, code v0.7):
\begin{itemize}
\item Qwen 2.5-0.5B-Instruct (4-bit, MLX), 24 layers, 896 hidden, 151K vocab
\item SmolLM2-360M-Instruct, 32 layers, 960 hidden, 49K vocab
\item SmolLM2-135M-Instruct, 30 layers, 576 hidden, 49K vocab
\end{itemize}

All measurements on raw mode: single token input, no template, no baseline subtraction. 5,000 tokens randomly sampled from each vocabulary.

\subsection{PCA Decomposition}
Full SVD at every layer: $\mathbf{X} = \mathbf{U}\mathbf{S}\mathbf{V}^T$ where $\mathbf{X} \in \mathbb{R}^{N \times d}$ is the centered exit state matrix. First 50 components retained. Stored as \texttt{decomp/L\{i\}\_scores.npy} in the MRI directory.

\subsection{Intrinsic Dimensionality}
$k$-NN maximum likelihood estimator (Levina--Bickel, 2004):
\[
\hat{d} = \frac{k-1}{\frac{1}{N}\sum_{i=1}^{N} \log\frac{r_k(x_i)}{r_1(x_i)}}
\]
where $r_j(x_i)$ is the distance from $x_i$ to its $j$-th nearest neighbor. $k=10$, cosine distance. Computed on the full-dimensional exit states, not the PCA projection.

\subsection{Neighbor Stability}
For each token, compute its $k=10$ nearest neighbors at layer $l$ and layer $l-1$. Stability = mean Jaccard overlap of the neighbor sets across all tokens.

\subsection{UMAP Script Purity}
UMAP projection to 3D ($n_{\text{neighbors}}=15$, $\min_{\text{dist}}=0.1$). For each token, compute the fraction of its 10 nearest neighbors (in UMAP space) that share the same Unicode script classification.

\section{Results}

\subsection{PCA Reports Crystallization}

\begin{table}[h]
\centering
\begin{tabular}{lcccc}
\toprule
Model & Crystal Layer & PC1 at Crystal & PC1 in Frozen & PC1 at Exit \\
\midrule
Qwen 0.5B & L2 & 97.9\% & 98.1\% & 44.0\% \\
SmolLM 360M & L3 & 98.6\% & 98.6\% & 67.1\% \\
SmolLM 135M & L11 & 99.7\% & 99.7\% & 98.0\% \\
\bottomrule
\end{tabular}
\caption{PCA PC1 variance ratio by depth region. All three models show apparent collapse to 1 dimension in the frozen zone.}
\end{table}

\subsection{Intrinsic Dimensionality Contradicts PCA}

\begin{table}[h]
\centering
\begin{tabular}{lccccc}
\toprule
Layer Region & PCA PC1 & Intrinsic Dim & Neighbor Stab. & UMAP Purity \\
\midrule
L0 (pre-crystal) & 4--10\% & 28--46 & --- & --- \\
Crystal birth & 97--99\% & 31--40 & 0.31--0.42 & --- \\
Frozen zone & 98--99\% & 34--37 & 0.93 & 0.655 \\
Output assembly & 44--76\% & 31--33 & 0.57--0.75 & 0.777 \\
\bottomrule
\end{tabular}
\caption{Qwen 0.5B. PCA reports 1 dimension in the frozen zone. $k$-NN MLE reports 34--37 dimensions. The discrepancy is a factor of 35$\times$.}
\end{table}

\finding{The frozen zone has 34--37 intrinsic dimensions at every layer, while PCA reports 98.6\% on one axis.}

\subsection{Neighbor Stability Shows Structure Preservation}

The 93\% neighbor stability in the frozen zone means that 9.3 of each token's 10 nearest neighbors remain the same between adjacent layers. The tokens are not shuffled randomly onto the dominant axis --- they maintain their relative arrangement. This arrangement has 35 meaningful dimensions that PCA projects to zero.

At the crystal birth layer, only 31--42\% of neighbors survive --- the representation is reorganized. At the output assembly (L21 in Qwen), only 57\% survive --- another reorganization.

\subsection{UMAP Reveals Language Clusters}

In the frozen zone (L10, Qwen 0.5B), UMAP script purity is 0.655 vs. 0.44 random baseline. Tokens cluster by Unicode script --- CJK near CJK, Latin near Latin, Arabic near Arabic --- on a manifold that PCA projects to a single line.

At the output assembly (L21), purity increases to 0.777. The model separates scripts \textit{more} as it prepares for output.

\subsection{Individual Trajectories Zigzag}

Token trajectories through depth, visualized as (PC0, PC1, layer), reveal oscillatory behavior in the frozen zone. Individual tokens cross back and forth on the dominant axes 3--5 times between L3--L20, while the aggregate PC1 ratio remains at 98.6\%.

This is consistent with the ``constant-bias'' gate neurons discovered in Session 6 being constant in aggregate but alternating in direction --- different layers push tokens in opposite directions along PC1. The net effect cancels; the per-token effect is a zigzag.

\open{Is the zigzag real oscillation or a projection artifact? Showing the same trajectory on multiple PC pairs simultaneously would distinguish these. If the zigzag appears on all pairs, the token is genuinely oscillating in the full space.}

\section{Discussion}

PCA variance ratios have been widely used as a measure of representational dimensionality in transformer interpretability research. Our results show this measure can be misleading by a factor of 35$\times$: a representation with 35 intrinsic dimensions is reported as having 1.

The frozen zone is not idle. It maintains a 35-dimensional manifold with language-specific clustering, reorganizes individual token positions through oscillatory trajectories, and preserves 93\% of local neighborhood structure between layers. This computation is invisible to any analysis that relies on PCA's dominant components.

The practical implication: layer pruning studies that identify ``skippable layers'' based on low delta norms or high PC1 ratios may be removing layers that perform essential computation in the minor dimensions.

\section{Methodology Note}

All MRI data recaptured in this session with verified decomposed forward (decomposed exit states match fused forward to float16 precision). The gradient computation uses \texttt{mx.stop\_gradient} instead of the numpy round-trip from Session 6, verified against finite differences (mean ratio 0.997$\pm$0.010). All measurements reproducible via \texttt{heinrich profile-*} commands with \texttt{--json} output and signal recording to the database.

\end{document}
