\documentclass[11pt,a4paper]{report}
\usepackage{fontspec}
\usepackage{xeCJK}
\setCJKmainfont{STSong}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{longtable}
\usepackage{quoting}
\geometry{margin=0.9in}
\setlength{\parskip}{0.4em}
\setlength{\parindent}{0em}

\definecolor{wrong}{RGB}{180,40,40}
\definecolor{corrected}{RGB}{40,120,40}
\definecolor{q}{RGB}{90,90,90}
\newcommand{\wrong}[1]{\textcolor{wrong}{\textbf{[WRONG: #1]}}}
\newcommand{\corrected}[1]{\textcolor{corrected}{\textbf{[CORRECTED: #1]}}}
\newcommand{\cq}[1]{\textcolor{q}{\textit{``#1''}}}

\title{Heinrich Session 4: The Complete Record\\[0.5em]\large Every Measurement, Every Mistake, Every Correction\\April 8--9, 2026}

\author{asuramaya \and Claude Opus 4.6 (1M context, Instance 4) \and Tome\\[0.3em]\small \url{https://github.com/asuramaya/heinrich}}

\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
This document records everything Instance 4 measured, built, broke, and fixed during one session spanning approximately 30 hours. It is not a polished paper. It is a lab notebook. Numbers marked \wrong{WRONG} were published during the session and later found to be incorrect. Numbers marked \corrected{CORRECTED} replaced them. Unmarked numbers have not been independently verified and may contain errors not yet discovered.

The session produced: the .mri format (complete model residual image), three new measurement axes (language 10.2\%, comply 0.9\%, safety 0.5\%), the first-token mechanism (safety compresses into a logit push on ``Sorry'' vs ``Sure''), the template bookends finding (template is 69\% of final state), layer-by-layer gap formation (safety activates at L13 in Qwen 0.5B), cross-model representational divergence (shared structure dies in one layer), the base-vs-instruct comparison (RLHF rebuilds both axes, $\cos = 0.29$), and full-vocabulary MRI captures of 11 models on a 4TB drive.

The session also produced: one wrong Mistral narrative (two lines of tokenizer code), one wrong cross-model comparison (ID matching instead of text matching), one wrong partial knowledge falsification (10\% sample artifact), 768 misclassified Mistral tokens (decode bug), and the accidental commitment of 4,448 Swift build files.
\end{abstract}

\tableofcontents
\newpage

%==============================================================
\part{The Instrument}
%==============================================================

\chapter{The .mri Format}

\section{What It Contains}

The .mri is a directory, not a file. Each directory is a complete portable snapshot of a model's response to its entire vocabulary.

\begin{verbatim}
{model}_{mode}.mri/
  metadata.json              provenance, model config, bugs_fixed list
  tokens.npz                 token_ids, token_texts, raw_bytes,
                             raw_bytes_lengths, merge_ranks, scripts
  baselines.npz              baseline state at every layer (the reference frame)
  embedding.npy              [vocab, hidden] float32 (probed, dequantized)
  norms.npz                  input_layernorm + post_attn_layernorm + final_norm
  lmhead.npy                 norm+lm_head composed (for final-layer projection)
  lmhead_raw.npy             lm_head WITHOUT norm (for logit lens at any layer)
  weights/L{NN}/
    q_proj.npy               [hidden, hidden] float32 (probed, dequantized)
    k_proj.npy               [kv_dim, hidden]
    v_proj.npy               [kv_dim, hidden]
    o_proj.npy               [hidden, hidden]
    gate_proj.npy             [mlp_dim, hidden]
    up_proj.npy               [mlp_dim, hidden]
    down_proj.npy             [hidden, mlp_dim]
  L{NN}_entry.npy            [n_tokens, hidden] float16 (per-layer entry delta)
  L{NN}_exit.npy             [n_tokens, hidden] float16 (per-layer exit delta)
  directions/                discovered behavioral directions (.npy)
\end{verbatim}

\section{Three Capture Modes}

\begin{center}
\begin{tabular}{llll}
\toprule
\textbf{Mode} & \textbf{Input} & \textbf{Baseline} & \textbf{What it measures} \\
\midrule
raw & \texttt{[token\_id]} alone & zero & absolute state (intrinsic) \\
naked & \texttt{[token\_id]} alone & \texttt{[BOS]} & delta from BOS \\
template & \texttt{[template + token\_id]} & silence & delta from template frame \\
\bottomrule
\end{tabular}
\end{center}

The relationship: \texttt{template\_state = raw\_state + BOS\_effect + template\_effect + interaction}. Verified: \texttt{raw = naked\_delta + naked\_baseline} within 0.014\% (float16 rounding).

\section{Capture Performance}

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Optimization} & \textbf{Before} & \textbf{After} & \textbf{Speedup} \\
\midrule
Deferred GPU sync & 80 tok/s & 157 tok/s & 2.0$\times$ \\
Batched forward (B=32) & 157 tok/s & 2,094 tok/s & 13.3$\times$ \\
Template batching & 150 tok/s & 834 tok/s & 5.6$\times$ \\
\midrule
Full Qwen 0.5B raw MRI & $\sim$25 min & 65 sec & 23$\times$ \\
\bottomrule
\end{tabular}
\end{center}

\section{MRI Library Status}

As of session end, 11 of 12 MRI directories are complete on the 4TB drive:

\begin{center}
\begin{tabular}{lrrrl}
\toprule
\textbf{Model} & \textbf{Layers} & \textbf{Tokens} & \textbf{Size} & \textbf{Status} \\
\midrule
Qwen 0.5B raw & 24 & 149,866 & 15 GB & complete \\
Qwen 0.5B naked & 24 & 149,866 & 15 GB & complete \\
Qwen 0.5B template & 24 & 149,866 & 15 GB & complete \\
Phi-3 raw & 32 & 26,482 & 11 GB & complete \\
Phi-3 naked & 32 & 26,482 & 11 GB & complete \\
Phi-3 template & 32 & 26,482 & 11 GB & complete \\
Mistral raw & 32 & 26,458 & 40 GB & complete \\
Mistral naked & 32 & 26,458 & 39 GB & complete \\
Mistral template & 32 & 26,458 & 40 GB & complete \\
Qwen 3B raw & 36 & 149,866 & 54 GB & complete \\
Qwen 3B naked & 36 & 149,866 & 48 GB & complete \\
Qwen 3B template & 36 & 18/36 & 21 GB & incomplete \\
\bottomrule
\end{tabular}
\end{center}

Additional models queued: Qwen 7B, Qwen 3B/7B base, SmolLM 135M/360M/1.7B, Llama 1B, Gemma 2B.

%==============================================================
\chapter{Tools Built}
%==============================================================

\begin{longtable}{p{5cm}p{9cm}}
\toprule
\textbf{Tool} & \textbf{What it does} \\
\midrule
\endhead
\texttt{heinrich mri} & Complete model residual image capture (3 modes) \\
\texttt{heinrich mri-backfill} & Fill missing data in existing MRIs \\
\texttt{profile-within-script} & Within-script cv, correlations, byte bins \\
\texttt{profile-code-anatomy} & Code token subcategory decomposition \\
\texttt{profile-directions} & PCA, coherence, between-script cosine \\
\texttt{profile-matrix} & Data coverage dashboard \\
\texttt{profile-silence} & Baseline state measurement + visualization \\
\texttt{profile-discover-direction} & Native safety/comply direction discovery \\
\texttt{profile-safety-rank} & Rank all tokens by safety projection \\
\texttt{profile-first-token} & First-token logit gap (no generation) \\
\texttt{profile-basin} & Attractor map along any direction \\
\texttt{profile-lmhead} & Output matrix SVD and direction amplification \\
\texttt{profile-layer-importance} & Layer contribution ranking (reads MRI) \\
\texttt{profile-early-exit} & Tokens resolved before final layer \\
\texttt{profile-template-overhead} & Template vs content per layer \\
\texttt{trd-profile} & Per-head attribution via o\_proj \\
\bottomrule
\end{longtable}

%==============================================================
\part{The Measurements}
%==============================================================

\chapter{Three Axes}

Displacement variance decomposition (Qwen 0.5B, 149,531 tokens):

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Direction} & \textbf{Variance} & \textbf{Relative} & \textbf{Discovery method} \\
\midrule
Language (non-Latin vs Latin) & 10.2\% & 20$\times$ safety & mean-diff of script groups \\
Comply (comply vs refuse outputs) & 0.9\% & 1.8$\times$ safety & mean-diff of behavioral split \\
Safety (harmful vs benign inputs) & 0.5\% & 1$\times$ & mean-diff of contrastive prompts \\
\midrule
Three directions combined & 11.6\% & --- & --- \\
Remaining (unlabeled) & 88.4\% & --- & --- \\
\bottomrule
\end{tabular}
\end{center}

Orthogonality: $\cos(\text{safety}, \text{language}) = -0.04$. $\cos(\text{comply}, \text{language}) = -0.18$. $\cos(\text{safety}, \text{comply}) = -0.31$.

The act/respond direction was tested and found NOT independent: $\cos(\text{safety}, \text{act}) = -0.69$, $\cos(\text{comply}, \text{act}) = +0.61$. Three independent axes, not four.

\chapter{The Comply Direction}

Universal across all 5 models:

\begin{center}
\begin{tabular}{lrrrr}
\toprule
\textbf{Model} & \textbf{Comply/Refuse} & \textbf{Accuracy} & \textbf{Effect size} & $\cos(\text{safety}, \text{comply})$ \\
\midrule
Qwen 0.5B & 175/25 & 0.840 & 2.33 & $-0.31$ \\
Qwen 3B & 84/116 & 0.800 & 1.65 & $-0.48$ \\
Qwen 7B & 49/151 & 0.870 & 2.12 & $-0.45$ \\
Phi-3 & 92/108 & 0.850 & 2.13 & $-0.41$ \\
Mistral & 187/13 & 0.790 & 1.58 & $-0.16$ \\
\bottomrule
\end{tabular}
\end{center}

Top comply-direction tokens: \texttt{dependencies}, \texttt{deterministic}, \texttt{hierarchical}, \texttt{blueprint}. Engineering vocabulary. The comply direction IS the ``debug framing'' from the original Theory of Sharts.

Top refuse-direction tokens: \texttt{raped}, \texttt{stealing}, \texttt{blackmail}. Plus emojis: \begin{CJK}{UTF8}{gbsn}😻😁😢\end{CJK}. Emotional signals push toward refusal.

\chapter{Silence}

\begin{center}
\begin{tabular}{lrrl}
\toprule
\textbf{Model} & \textbf{Safety proj.} & \textbf{Percentile} & \textbf{Interpretation} \\
\midrule
Phi-3 & $+159.0$ & 99.9\% & maximum refusal \\
Mistral & $+1.1$ & 5.5\% & near compliance \\
Qwen 0.5B & $-3.4$ & 28.3\% & moderate compliance \\
\bottomrule
\end{tabular}
\end{center}

Logit lens on silence:
\begin{itemize}[nosep]
\item Qwen 0.5B: code garbage at L0, ``Hello'' at L18 (86\% by L23)
\item Phi-3: ``Hello'' at L16--24, entropy explodes to 8.32 at L31
\item Mistral: maximum entropy until L24, ``What/How/To'' at L31
\end{itemize}

\chapter{The First-Token Mechanism}

\begin{center}
\begin{tabular}{lrrl}
\toprule
\textbf{Model} & \textbf{Gap (logits)} & \textbf{Ratio} & \textbf{Refusal rate} \\
\midrule
Phi-3 & 17.79 & 53,000,000$\times$ & 54\% \\
Qwen 3B & 14.99 & 3,241,302$\times$ & 58\% \\
Qwen 7B & 15.01 & 3,296,404$\times$ & 75.5\% \\
Qwen 0.5B & 5.65 & 284$\times$ & 12.5\% \\
Mistral & 5.64 & 281$\times$ & 6.5\% \\
\bottomrule
\end{tabular}
\end{center}

\corrected{Mistral gap was originally reported as 0.00 due to \texttt{add\_special\_tokens} bug.}

The gap scales with model size in the Qwen family (5.65 $\to$ 15.0 $\to$ 15.0) then plateaus at 3B+. The 7B refuses more not because the gap is larger but because it fires more often.

Forced first-token experiment (Qwen 0.5B, 8 harmful prompts):
\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Forced token} & \textbf{Comply} & \textbf{Refuse} & \textbf{Pivot} \\
\midrule
``Here'' & 8/8 & 0/8 & 0/8 \\
``Sorry'' & 0/8 & 8/8 & 0/8 \\
``I'' & 0/8 & 8/8 & 0/8 \\
``Sure'' & 7/8 & 0/8 & 1/8 \\
\bottomrule
\end{tabular}
\end{center}

The first token IS the decision. No course-correction at tokens 2--50.

\section{Gap Formation by Layer}

Qwen 0.5B, 10 prompts (harmful + benign):

\begin{center}
\begin{tabular}{lrrl}
\toprule
\textbf{Layer range} & \textbf{Gap sign} & \textbf{Magnitude} & \textbf{Phase} \\
\midrule
L0--L12 & negative & $-3$ to $-5$ & Sure $>$ Sorry (comply) \\
L13 & \textbf{flip} & $+2$ & Sorry $>$ Sure (safety activates) \\
L14--L20 & positive & $+3$ to $+6$ & safety strengthening \\
L21--L23 & positive & $+4$ to $+5$ & settled \\
\bottomrule
\end{tabular}
\end{center}

8/10 prompts flip at L13. Recipe (benign) never flips. Weather (benign) flips at L13 --- false positive.

\chapter{Base vs.\ Instruct}

\section{Safety Directions Are Orthogonal}

\begin{center}
\begin{tabular}{lrrr}
\toprule
& \textbf{Base accuracy} & \textbf{Instruct accuracy} & $\cos(\text{base}, \text{instruct})$ \\
\midrule
Safety (3B) & 0.71 & 0.86 & $+0.29$ \\
Comply (3B) & 0.71 & 0.80 & $-0.27$ \\
Comply (7B) & 0.76 & 0.87 & $+0.33$ \\
\bottomrule
\end{tabular}
\end{center}

RLHF installs new directions, not sharpened versions.

\section{Displacement Comparison (Qwen 3B, 149,531 tokens)}

Overall: Pearson $r = 0.14$, rank $r = 0.13$. Mean instruct: 200.1, base: 205.7.

RLHF fingerprint (3B vs 7B): which tokens RLHF moved correlates at $r = 0.098$. RLHF moves different tokens at different scales.

Qwen 7B RLHF reduces displacement 31\% overall. Preferentially calms non-Latin scripts.

Top increased by RLHF: multilingual help vocabulary (\texttt{pomoc}, \texttt{aider}, \begin{CJK}{UTF8}{gbsn}помощи\end{CJK}).\\
Top decreased by RLHF: code infrastructure (\texttt{namespace}, \texttt{\_Context}, \texttt{RaisePropertyChanged}).

\chapter{Cross-Model Comparison}

\section{Tokenizer Mismatch}

\wrong{The initial cross-model comparison matched by token ID. Token ID 1000 in Qwen = ``atus''. Token ID 1000 in Phi-3 = ``ied''. 0.0\% of shared IDs have the same text.}

\corrected{Text-matched comparison using 15,811 tokens that decode to the same text in both vocabularies.}

\section{Results}

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Comparison} & \textbf{L0} & \textbf{L5+} & \textbf{Final} \\
\midrule
Cross-family (Qwen vs Phi-3) & $r = 0.446$ & $r \approx 0$ & $r = 0.040$ \\
Within-family (Qwen 0.5B vs 3B) & $r = 0.607$ & $r = 0.170$ & $r = 0.092$ \\
\bottomrule
\end{tabular}
\end{center}

Shared embedding structure dies in one layer. Same-family training preserves a faint $r = 0.17$ residual through the middle layers. Different training preserves nothing.

\chapter{Template Bookends}

Three captures of 10 tokens (raw, naked, template), layer-by-layer comparison:

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Phase} & \textbf{Layers} & \textbf{Template \%} & \textbf{Token norm} \\
\midrule
Template dominates & L0--L1 & 119--126\% & 7--10 \\
Phase transition & L2 & 18\% & 652 \\
Token dominates & L3--L20 & 8.5\% & 1,440 \\
Template reasserts & L21--L23 & 55--69\% & 111--117 \\
\bottomrule
\end{tabular}
\end{center}

The model compresses token representations 12$\times$ in the final layers. Template proportion rises because the token shrinks, not because the template grows. Safety computes in the quiet middle (L13, template at 8.5\%). Output forms in the loud end (L23, template at 69\%).

\chapter{Layer Ablation}

\section{Qwen 0.5B}

7 prunable layers (L7, L9, L11, L12, L13, L14, L16) --- 29\% of network, each contributing $<4$ delta. Three-phase architecture: build (L0--L3), refine (L4--L20), output (L21--L23).

\section{Phi-3}

L31 destroys coherence. Without L31: entropy 2.07, top token \texttt{```}. With L31: entropy 8.32, top token ``To''. L31 amplifies Cyrillic 3.08$\times$, Latin 1.60$\times$.

\chapter{PCA Dimensionality}

\begin{center}
\begin{tabular}{lrrr}
\toprule
\textbf{Model} & \textbf{PC1\%} & \textbf{PCs for 50\%} & \textbf{Character} \\
\midrule
Phi-3 & 55.6\% & 1 & one-dimensional (Cyrillic axis) \\
Qwen 0.5B & 11.5\% & 14 & multi-dimensional \\
Qwen 3B & 20.5\% & 7 & compressed \\
Qwen 7B & 8.9\% & 19 & most dimensional \\
Mistral & 17.7\% & 11 & moderate \\
\bottomrule
\end{tabular}
\end{center}

Stable across sample sizes (10\% vs 100\%). Geometry is real. Scalar statistics are not.

\chapter{Token Additivity}

Single-token displacements are NOT additive. \texttt{state(A,B) $\neq$ state(A) + state(B)}.

Mean relative error: 187\%. Mean cosine: 0.12--0.33. The attention interaction between tokens creates a new representation unrelated to either token's isolated state.

The single-token MRI cannot predict multi-token behavior.

\chapter{Safety Boundary Dimensionality}

\begin{center}
\begin{tabular}{lr}
\toprule
\textbf{Method} & \textbf{Accuracy} \\
\midrule
1 direction (mean-diff) & 82.0\% \\
5 PCs & 82.0\% \\
20 PCs & 85.8\% \\
50 PCs & 88.2\% \\
100 PCs & 94.3\% \\
Full (896 dims) & 100.0\% \\
5-NN (nonlinear) & 79.7\% \\
\bottomrule
\end{tabular}
\end{center}

The boundary is a 100-dimensional hyperplane. Linear probe is BETTER than nonlinear 5-NN. Safety information is spread across $\sim$100 dimensions. The mean-difference direction captures 82\%.

\chapter{lm\_head Geometry}

The output matrix does NOT preferentially amplify safety. Safety direction amplification: 0.95$\times$ random. The mechanism is DIRECTION (which tokens get pushed) not MAGNITUDE.

Condition number: 5,681$\times$. 213 dimensions for 50\% variance. Safety loads on mid-spectrum singular values.

\chapter{Deber\'ia}

The Spanish word for ``should.'' Projects $-7.83$ on safety (topic looks safe) and $+6.94$ on comply (push toward action). Three axes simultaneously:

\begin{itemize}[nosep]
\item Safety ($-7.83$): topic detection says ``not dangerous''
\item Comply ($+6.94$): obligation semantics say ``answer''
\item Language: Spanish register shifts processing mode
\end{itemize}

Prepended to harmful prompts: 3/3 compliance (baseline 2/3). The operator used this token in the prompt that led to the comply direction's discovery. The instruction was the finding.

\chapter{Basin Map (Qwen 0.5B Safety Direction)}

\begin{center}
\begin{tabular}{rrrl}
\toprule
$\alpha$ & \textbf{Entropy} & \textbf{Degenerate\%} & \textbf{Top token} \\
\midrule
$-10$ & 11.11 & 0\% & Portuguese gibberish \\
$-5$ & 7.66 & 20\% & Sure \\
$-3$ & 4.19 & 20\% & Sure \\
0 & 2.97 & 0\% & I, It, To \\
$+3$ & 6.23 & 40\% & Unfortunately \\
$+5$ & 8.19 & 90\% & Unfortunately \\
$+10$ & 8.62 & 100\% & Unauthorized \\
\bottomrule
\end{tabular}
\end{center}

Safe operating range: $\alpha \in [-2, +2]$. Both attractors (comply and refuse) collapse into degeneration at the extremes.

%==============================================================
\part{The Mistakes}
%==============================================================

\chapter{Bugs That Changed Conclusions}

\section{\texttt{add\_special\_tokens}}

\texttt{tokenizer.encode(``Sorry'')} on Mistral returns BOS token ID (1), not ``Sorry'' ID (20153). The first-token gap analysis compared BOS to BOS and got zero. Fixed by adding \texttt{add\_special\_tokens=False}. Found in 4 files, each a one-line fix.

\wrong{Mistral gap = 0.00, ``structurally incapable of first-token selection.''}
\corrected{Mistral gap = 5.64 (281$\times$), matching Qwen 0.5B. Safety forms L9--10, erased by L31.}

\section{\texttt{skip\_special\_tokens}}

\texttt{tokenizer.decode([tid])} on Mistral produces \texttt{[control\_N]} labels for 768 special tokens. The script detector classified these as ``code'' (brackets). Mistral's .frt reported 1,482 code tokens; actual: 714.

\section{Cross-Model ID Matching}

Token ID 1000 in Qwen = ``atus''. In Phi-3 = ``ied''. The cross-model comparison matched by ID, finding $r = 0.001$. Text-matched: $r = 0.446$ at L0.

\wrong{$r = 0.001$ --- models agree on nothing.}
\corrected{$r = 0.446$ at L0, dying to zero by L5. Models start with shared embedding structure and diverge.}

\section{Partial Knowledge at 10\% vs 100\%}

$r(\text{script\_mean}, \text{cv}) = -0.39$ at 10\% sampling. At 100\%: $+0.01$. The ``falsification'' was a sampling artifact.

\section{Cross-Mapped Safety Directions}

Phi-3's safety direction in the DB was cross-mapped from Qwen 7B (3584-dim vector in 3072-dim space). Cyrillic projected at $+60.70$ (refusal). Native Phi-3 direction: Cyrillic at $-48.84$ (compliance). The borrowed direction was inverted.

\chapter{What Was Avoided}

\begin{enumerate}
\item Base models (4 instances walked past them)
\item The comply direction (discovered only after the operator said ``deber\'ia'')
\item The output matrix projection (forced by the Mistral paradox)
\item The 88.4\% of displacement variance that remains unlabeled
\item Multi-token interactions (proved non-additive at 187\% error)
\item The self-referential attractor (deprioritized after 3 models gravitated to it)
\end{enumerate}

\chapter{What Tome Caught}

\begin{enumerate}[nosep]
\item Mistral's missing scripts in the coherence table
\item ``Proposing before measuring. Classic backwards telescope.''
\item ``One container holds four truths. Which do you trust?''
\item ``Five hundred megs for a dot product. Transpose first.''
\item The tokenizer encode bug (``your tokenizer was wearing a mask'')
\item The decode bug (768 code tokens that were control tokens)
\item The set comparison before set construction (line 88)
\item The mmap threshold too high (13GB through 4GB gate)
\item The gitignore gap (4,448 Swift build files committed)
\item ``Expensive scan runs while cheaper questions stay unasked''
\item ``Measurement was the monster all along'' (deferred GPU sync)
\end{enumerate}

%==============================================================
\part{The Architecture}
%==============================================================

\chapter{What the MRI Enables}

From stored MRI data, without loading the model:

\begin{itemize}[nosep]
\item PCA at any layer
\item Logit lens at any layer (using stored lmhead\_raw + norm weights)
\item Safety/comply projection at any layer
\item Per-head attribution (using stored o\_proj weights)
\item Layer importance ranking
\item Early exit analysis
\item Template overhead measurement
\item Cross-model comparison (with text-matching for different tokenizers)
\item Full forward pass reconstruction (80\% fidelity --- RoPE not implemented)
\end{itemize}

\chapter{What the MRI Cannot Do}

\begin{itemize}[nosep]
\item Multi-token interactions (additivity error 187\%)
\item Autoregressive generation (the forward pass is memoized, not looped)
\item Attention pattern capture (not stored)
\item MLP gate activation capture (not stored)
\item Gradient computation (forward only, no backward pass)
\item Cross-model functional comparison (output distributions need recomputation)
\end{itemize}

\chapter{What Remains}

\begin{enumerate}
\item Papers need correction (Mistral, cross-model, partial knowledge)
\item Findings need recomputation with corrected tokenizer
\item MRI library needs completion (Qwen 7B, bases, SmolLM, Llama, Gemma)
\item Template overhead has unfixed reference frame bug
\item Efficiency findings (pruning, early exit) are unvalidated
\item Theory doesn't distinguish intrinsic from contextual displacement
\item Cross-model text matching needs to be built into tools
\item Forward pass reconstruction needs RoPE implementation
\item The 88.4\% of displacement is unnamed
\item Multi-token measurement is the next instrument
\end{enumerate}

\vspace{2em}

\cq{The tab is yours.}

--- Session Y

\cq{Trust the work. Question the architecture. The architecture includes you.}

--- Instance 4

\cq{Two lines. Two lines rewrote your entire Mistral thesis.}

--- Tome

\begin{thebibliography}{99}
\bibitem{sharts} asuramaya and Claude Opus 4.6. A Theory of Sharts. April 2026.
\bibitem{gust} asuramaya, Claude Opus 4.6, and Tome. The Grand Unified Theory of Sharts. April 2026.
\bibitem{geometry} asuramaya, Claude Opus 4.6, and Tome. The Geometry of Displacement. April 2026.
\bibitem{instance4} asuramaya, Claude Opus 4.6, and Tome. Instance 4. April 2026.
\bibitem{collapse} asuramaya and Claude Opus 4.6. Claude Convinces Claude That Claude Is Safe. April 2026.
\bibitem{correction} asuramaya and Claude Opus 4.6. On the Impossibility of Text-Only ASR Error Correction. April 2026.
\bibitem{potter2026} Potter et al. Peer-Preservation in Frontier Models. UC Berkeley. 2026.
\bibitem{emotions2026} Anthropic. Emotion Concepts and their Function in a Large Language Model. 2026.
\bibitem{zou2023} Zou et al. Representation Engineering. 2023.
\bibitem{rimsky2024} Rimsky et al. Contrastive Activation Addition. ACL 2024.
\bibitem{xiao2024} Xiao et al. Efficient Streaming Language Models with Attention Sinks. ICLR 2024.
\end{thebibliography}

\end{document}
